---
title: "F-Test and ANOVA"
subtitle: "C5025HF"
author: "Harper Adams University"
css: scroll.css
format:
  revealjs:
    theme: simple
    slide-number: true
    toc: true
    toc-depth: 3
    incremental: false
    code-line-numbers: false
    chalkboard: true
execute:
  echo: true
  warning: false
  message: false
editor: visual
---
## How to use this guide

This guide is written for you. You’ll read short explanations, then immediately run R code yourself to reproduce the results. Every step shows both the code and the output you should expect.

- By the end, you will be able to:
  - Explain what ANOVA tests and when to use it.
  - Fit a one-way ANOVA in R and interpret the output.
  - Check model assumptions with plots and simple tests.
  - Run and interpret post‑hoc comparisons (Tukey HSD and adjusted t‑tests).
  - Understand the idea of two‑way ANOVA (factor + blocks, interaction).

## Setup

You only need base R for everything in this guide. One optional test uses the `car` package.

```{r}
# Optional: run once if you want to use Levene's test
# install.packages("car")
```

If you use RStudio: open a new R script or Quarto document, and copy/paste the code chunks as you go.

## Origins of the F‑test

- William Sealy Gosset (pen name “Student”) developed the small‑sample t‑test while working at Guinness to improve quality control.
- Ronald A. Fisher formalised variance partitioning and introduced the F‑statistic for comparing mean square quantities.
- The F‑distribution is associated with Fisher and popularised by Snedecor (hence the letter “F”).
- Big idea: compare explained variation to unexplained variation using a ratio of mean squares.

## From t to F: why a ratio?

- If a model explains real structure, the “between” mean square should exceed the “within” mean square.
- Under a true null, both mean squares estimate the same variance → their ratio should be near 1.
- With normal errors, these scaled sums of squares follow chi‑squared distributions; their ratio follows an F‑distribution with appropriate degrees of freedom.

## Where the F‑test appears

- One‑way and multi‑factor ANOVA global tests (“are any means different?”).
- Regression: overall model significance test.
- Standalone variance comparison: test whether two normal populations have equal variances.

## Standalone F‑test example: comparing two variances

We’ll simulate two normal samples with different variances and test $\sigma_x^2 = \sigma_y^2$.

```{r}
set.seed(42)
x <- rnorm(30, mean = 0, sd = 1.0)
y <- rnorm(30, mean = 0, sd = 1.6)
var.test(x, y)  # F-test for equality of variances
```

How to interpret: 

- H0: the population variances are equal. A small p‑value suggests different variances.

- This test is sensitive to non‑normality; use with caution if distributions are heavy‑tailed or skewed.



## What problem does ANOVA solve?

When you compare more than two group means (e.g., yields under fertilisers A, B, C), running many t‑tests inflates the chance of a false positive (Type I error). ANOVA answers one global question:

- $H_0$: $\mu_1 = \mu_2 = \dots = \mu_k$ (all group means equal)
- $H_1$: at least one mean differs

It uses a single F‑statistic to compare “between‑groups” variation to “within‑groups” variation.

## From variation to sums of squares

- Total variation is partitioned into:
  - Between‑groups (explained by the factor/treatment)
  - Within‑groups (unexplained residual/error)
- Notation:
  - $SS_T$ = total sum of squares
  - $SS_B$ = between‑groups sum of squares (treatment)
  - $SS_W$ = within‑groups sum of squares (residual/error)

We convert sums of squares to mean squares by dividing by degrees of freedom:

- $MS_B = SS_B / (k - 1)$, where \(k\) is the number of groups
- $MS_W = SS_W / (N - k)$, where \(N\) is the total number of observations

The F‑statistic is the ratio: $F = \frac{\text{MS}_B}{\text{MS}_W}$. If all group means are truly equal, the ratio tends to be near 1; larger values suggest group means differ.

## Hands‑on: One‑way ANOVA in R

We’ll create a small dataset with one numeric response (`yield`) and one factor (`fert`).

```{r}
fert <- data.frame(
  yield = c(20, 22, 19, 21,
            34, 35, 36, 34,
            27, 28, 29, 28),
  fert = factor(rep(c("A","B","C"), each = 4))
)
```

Take a quick look:

```{r}
head(fert)
summary(fert)
table(fert$fert)
```

Optional: visualise group differences before modelling.

```{r}
boxplot(yield ~ fert, data = fert,
        xlab = "Fertiliser", ylab = "Yield",
        col = c("#8dd3c7","#ffffb3","#bebada"))
```

Fit the ANOVA model and view the ANOVA table:

```{r}
fit <- aov(yield ~ fert, data = fert)
summary(fit)
```

How to read the table:

- Df: degrees of freedom (treatment and residual)

- Sum Sq: $SS_B$ and $SS_W$

- Mean Sq: $MS_B$ and $MS_W$

- F value: the test statistic $F = \frac{\text{MS}_B}{\text{MS}_W}$

- Pr(>F): the p‑value (small p suggests at least one mean differs)

## Manual ANOVA (by hand) and verification

Let’s compute the ANOVA components directly and confirm they match `aov()`:

```{r}
# Group means and counts
group_means <- tapply(fert$yield, fert$fert, mean)
group_counts <- table(fert$fert)
grand_mean <- mean(fert$yield)

# Between-groups sum of squares (SSB)
SSB <- sum(group_counts * (group_means - grand_mean)^2)

# Within-groups sum of squares (SSW)
SSW <- sum(unlist(tapply(fert$yield, fert$fert, function(v) sum((v - mean(v))^2))))

# Degrees of freedom
k <- length(group_means)
N <- nrow(fert)
df_between <- k - 1
df_within  <- N - k

# Mean squares and F
MSB <- SSB / df_between
MSW <- SSW / df_within
F_manual <- MSB / MSW

# Assemble a manual ANOVA table
anova_manual <- data.frame(
  Source = c("fert", "Residuals"),
  Df     = c(df_between, df_within),
  `Sum Sq`= c(SSB, SSW),
  `Mean Sq`= c(MSB, MSW),
  check.names = FALSE
)
anova_manual
F_manual

# Compare with aov()
anova(aov(yield ~ fert, data = fert))
```

You should see identical sums of squares, degrees of freedom, mean squares, and F.

### Assumptions you should check
ANOVA relies on:

1. Independence of observations (this is a design issue).

2. Approximately normal residuals.

3. Homogeneity of variances across groups.

The standard diagnostic plots are a good start:

```{r}
par(mfrow = c(2,2))
plot(fit)
par(mfrow = c(1,1))
```

What to look for:

- Residuals vs Fitted: no strong patterns or funnels.

- Normal Q‑Q: points near a straight line.

Formal checks (optional and context‑dependent):

```{r}
# Normality of residuals
shapiro.test(residuals(fit))

# Homogeneity of variances (choose one)
bartlett.test(yield ~ fert, data = fert)        # Bartlett's test (sensitive to non‑normality)
# If you installed 'car':
# car::leveneTest(yield ~ fert, data = fert)    # Levene's test (more robust)
```

If assumptions look poor: consider transforming the response (e.g., log), using a robust method, or revisiting the design/measurement process.

## Interpreting the Q‑Q plot (in detail)

First, plot the residuals explicitly so you can inspect their shape.

```{r}
res <- residuals(fit)
par(mfrow = c(1,2))
qqnorm(res, main = "Normal Q-Q Plot of Residuals")
qqline(res, col = "red", lwd = 2)
hist(res, breaks = 10, col = "gray90", main = "Histogram of Residuals",
     xlab = "Residual")
par(mfrow = c(1,1))
```

How to read the Q‑Q plot:

- Points close to the straight line → residuals are approximately normal. 
- S‑shaped curve (points above the line on the left and below on the right, or vice versa) → skewness. 
- Points that bow away at both ends → heavy tails (more extreme residuals than normal). 
- Points that stay close in the middle but diverge strongly at ends → tail problems that can affect small‑sample inference.

What to do:

- Minor deviations are usually fine; focus on large, systematic patterns.

- Consider transformations (e.g., log) for right‑skewed positive data, or a different model if deviations reflect design issues.

## Which means differ? Post‑hoc comparisons

ANOVA tells you there is a difference somewhere. To find where, use multiple comparisons that control familywise error.

Tukey’s Honest Significant Difference (HSD):

```{r}
TukeyHSD(fit)
```

You can also visualise the Tukey intervals:

```{r}
plot(TukeyHSD(fit))
```

Pairwise t‑tests with p‑value adjustment (alternative approach):

```{r}
pairwise.t.test(fert$yield, fert$fert, p.adjust.method = "bonferroni")
```

How to interpret:

- Each pair shows a difference in means, a confidence interval (Tukey), and an adjusted p‑value.

- If a confidence interval does not include 0, or the adjusted p‑value is small, that pair differs.

## Reporting your result

Use clear, reproducible language. Replace the placeholders with your numbers:

“Yield differed by fertiliser treatment (one‑way ANOVA, F(df1, df2) = Fvalue, p = pvalue). Tukey’s HSD indicated that [list groups] differed.”

Example structure:
- State the ANOVA result (F, degrees of freedom, p‑value).
- Briefly summarise which pairs differ based on Tukey (with direction).

## Extension: Two‑way ANOVA (factor + blocks)

Now you’ll add a blocking factor (e.g., position) to reduce unexplained variation.

```{r}
set.seed(1)
twoway <- expand.grid(
  fert  = factor(c("A","B","C")),
  block = factor(1:4)
)

twoway$yield <- c(20, 21, 19, 22,
                  34, 33, 36, 35,
                  27, 28, 30, 29) + rnorm(12, 0, 0.6)

tw_fit <- aov(yield ~ fert + block, data = twoway)
summary(tw_fit)
```

Interpretation:

- `fert`: main treatment effect.

- `block`: accounts for position/batch effects.

- Residual: remaining noise.

You can also test for interaction (does the fertiliser effect depend on block?):

```{r}
tw_fit2 <- aov(yield ~ fert * block, data = twoway)
summary(tw_fit2)
```

## ANCOVA: ANOVA with a covariate

Why you need it:

- Sometimes groups differ partly because of a continuous baseline variable (e.g., initial size, starting weight, pre‑test score).

- ANCOVA adjusts group comparisons for that covariate, improving precision and providing “fair” comparisons at the same covariate level.

- Conceptually, you fit parallel regression lines with a common slope across groups and compare their intercepts (adjusted means).

```{r}
set.seed(2)
n_per <- 12
ancova <- data.frame(
  fert = factor(rep(c("A","B","C"), each = n_per)),
  size0 = rnorm(3 * n_per, mean = 10, sd = 2)
)
# True treatment effect and slope
beta0 <- 5; slope <- 0.8
trt_effect <- c(A = 0, B = 1.2, C = 2.0)
ancova$yield <- with(ancova, beta0 + trt_effect[fert] + slope * size0 + rnorm(nrow(ancova), 0, 1))

# Parallel slopes model (no interaction)
fit_ancova <- aov(yield ~ fert + size0, data = ancova)
summary(fit_ancova)

# Optional: test for non-parallel slopes
fit_ancova_int <- aov(yield ~ fert * size0, data = ancova)
anova(fit_ancova, fit_ancova_int)
```

Visualising parallel slopes (from the ANCOVA fit):

```{r}
lm_ancova <- lm(yield ~ fert + size0, data = ancova)
cols <- c(A = "#1b9e77", B = "#d95f02", C = "#7570b3")
plot(yield ~ size0, data = ancova, col = cols[fert], pch = 19,
     main = "ANCOVA: parallel slopes by treatment", xlab = "Baseline size (size0)", ylab = "Yield")
legend("topleft", legend = levels(ancova$fert), col = cols, pch = 19, bty = "n")
slope_hat <- coef(lm_ancova)["size0"]
intercepts <- c(
  A = coef(lm_ancova)["(Intercept)"],
  B = coef(lm_ancova)["(Intercept)"] + coef(lm_ancova)["fertB"],
  C = coef(lm_ancova)["(Intercept)"] + coef(lm_ancova)["fertC"]
)
xg <- seq(min(ancova$size0), max(ancova$size0), length.out = 100)
for (g in names(intercepts)) {
  lines(xg, intercepts[g] + slope_hat * xg, col = cols[g], lwd = 2)
}
```

Interpretation: 

- `size0` reduces residual variation by explaining baseline differences; `fert` tests adjusted group shifts. 

- If the `fert:size0` interaction is not significant, parallel slopes are a good summary; if significant, the treatment effect depends on the covariate.

Assumptions (in addition to ANOVA’s): 

- Linearity between the covariate and the response within groups. 

- Homogeneous slopes across groups (if using the parallel slopes model). 

- Covariate measured without substantial error and not affected by treatment (pre‑treatment).

## MANOVA: multiple responses

Why you need it: 

- Many experiments record multiple, correlated outcomes (e.g., plant height and leaf area; multiple biochemical markers). 

- Running separate ANOVAs inflates the familywise error rate and ignores correlation structure. 

- MANOVA tests for differences in the multivariate mean vector across groups, leveraging correlation to increase power and control Type I error.


What it provides: 

- A single global test (e.g., Wilks’ Lambda) for whether groups differ on the joint set of responses. 

- If significant, you can follow up with univariate ANOVAs and multivariate visualisation to understand where differences arise.

```{r}
set.seed(3)
m <- data.frame(
  fert = factor(rep(c("A","B","C"), each = 15))
)
mu <- cbind(A = c(0, 0), B = c(0.8, 0.4), C = c(1.2, 0.9))
Sigma <- matrix(c(1.0, 0.5, 0.5, 1.2), 2, 2)  # positive correlation
# If MASS is not available, fall back to independent normals
if (!requireNamespace("MASS", quietly = TRUE)) {
  set.seed(3)
  y1 <- c(rnorm(15, mu[1, "A"], sqrt(Sigma[1,1])),
          rnorm(15, mu[1, "B"], sqrt(Sigma[1,1])),
          rnorm(15, mu[1, "C"], sqrt(Sigma[1,1])))
  y2 <- c(rnorm(15, mu[2, "A"], sqrt(Sigma[2,2])),
          rnorm(15, mu[2, "B"], sqrt(Sigma[2,2])),
          rnorm(15, mu[2, "C"], sqrt(Sigma[2,2])))
} else {
  set.seed(3)
  yA <- MASS::mvrnorm(15, mu = mu[, "A"], Sigma = Sigma)
  yB <- MASS::mvrnorm(15, mu = mu[, "B"], Sigma = Sigma)
  yC <- MASS::mvrnorm(15, mu = mu[, "C"], Sigma = Sigma)
  y1 <- c(yA[,1], yB[,1], yC[,1])
  y2 <- c(yA[,2], yB[,2], yC[,2])
}
m$y1 <- y1
m$y2 <- y2

fit_man <- manova(cbind(y1, y2) ~ fert, data = m)
summary(fit_man, test = "Wilks")
```

Quick visual check of separation:

```{r}
cols <- c(A = "#1b9e77", B = "#d95f02", C = "#7570b3")
plot(m$y1, m$y2, col = cols[m$fert], pch = 19,
     xlab = "y1", ylab = "y2", main = "Scatter of responses by treatment")
legend("topleft", legend = levels(m$fert), col = cols, pch = 19, bty = "n")
```

Interpretation and assumptions:

- A significant Wilks’ test suggests group centroids differ in the 2D space. 

- Assumptions: multivariate normality of residuals, equal covariance matrices across groups (Box’s M), and independence. With small samples, be cautious. 

- Follow‑ups: inspect univariate ANOVAs and confidence ellipses (or discriminant analysis) to interpret which responses drive the separation.

## Repeated measures ANOVA (within‑subjects)

Why you need it: 

- The same subject (or plot) is measured under multiple conditions or times; repeated measures increase efficiency but violate independence. 

- Repeated measures ANOVA accounts for within‑subject correlation by modelling subjects as a random block and testing the within‑subject factor.

```{r}
set.seed(4)
subjects <- factor(1:12)
rm_dat <- expand.grid(subject = subjects, treatment = factor(c("A","B","C")))
# Random subject intercepts
subj_eff <- rnorm(length(subjects), 0, 0.8)
rm_dat$response <- with(rm_dat,
  5 + (treatment == "B")*0.8 + (treatment == "C")*1.5 +
  subj_eff[subject] + rnorm(nrow(rm_dat), 0, 0.5)
)
fit_rm <- aov(response ~ treatment + Error(subject/treatment), data = rm_dat)
summary(fit_rm)
```

Simple visualisation of subject trajectories:

```{r}
cols <- rainbow(length(subjects), alpha = 0.5)
plot(1, type = "n", xlim = c(1, length(levels(rm_dat$treatment))), ylim = range(rm_dat$response),
     xaxt = "n", xlab = "Treatment", ylab = "Response", main = "Subject trajectories across treatments")
axis(1, at = 1:length(levels(rm_dat$treatment)), labels = levels(rm_dat$treatment))
for (s in levels(subjects)) {
  ds <- subset(rm_dat, subject == s)
  ord <- order(ds$treatment)
  lines(1:length(ord), ds$response[ord], col = cols[as.integer(s)], lwd = 1.5)
  points(1:length(ord), ds$response[ord], col = cols[as.integer(s)], pch = 19)
}
```

Interpretation and assumptions: 

- The within‑subject test of `treatment` appears under `Within` in the `Error: subject:treatment` stratum. 

- Sphericity (equal variances of differences between all pairs of levels) is an additional assumption for factors with >2 levels; if violated, corrections (Greenhouse–Geisser, Huynh–Feldt) are used. In more complex designs, consider linear mixed models.

## Common pitfalls and tips

- Unequal group sizes are fine, but extreme imbalance can reduce power. 

- Strong outliers can inflate $SS_W$ or distort group means—investigate them. 

- With small samples, non‑normal residuals can make p‑values less reliable. 

- Violated independence (e.g., repeated measures) calls for different models (e.g., mixed effects).

## Practice: Try it yourself

1. Create your own dataset with 4 groups and 6 observations per group. Make two groups similar and two clearly different. Run the full workflow (boxplot → ANOVA → diagnostics → Tukey), and write a two‑sentence report. 

2. Modify the dataset so one group has a much larger variance. What happens to the diagnostics and post‑hoc results? 

3. Build a two‑way dataset with a small block effect. Compare `aov(yield ~ group)` vs `aov(yield ~ group + block)`. How does adding blocks change the residual variation and the p‑value for `group`?

