---
title: "Correlation Analysis"
subtitle: "Direction, strength, assumptions, and interpretation"
author: "Harper Adams University"
format:
  revealjs:
    css: scroll.css
    theme: simple
    slide-number: true
    toc: true
    incremental: false
    chalkboard: true
execute:
  echo: true
  warning: false
  message: false
---

# Section 1: What correlation measures

## The question

Correlation describes the **direction** and **strength** of association between two variables.

- Positive association: larger values of one variable tend to accompany larger values of the other.
- Negative association: larger values of one variable tend to accompany smaller values of the other.
- Weak association: knowing one variable gives little information about the other.
- Strong association: observations follow a clear increasing or decreasing pattern.

## Start with a graph

```{r}
plot(
  iris$Sepal.Length,
  iris$Petal.Length,
  pch = 19,
  col = "steelblue",
  xlab = "Sepal length",
  ylab = "Petal length",
  main = "Sepal and petal length"
)
```

A correlation coefficient should always be interpreted alongside a graph because unusual shapes, clusters, and outliers can be hidden by a single number.

## Correlation and causation

An association does not by itself establish that one variable causes the other.

Possible explanations include:

- a direct causal effect;
- reverse causation;
- a third variable influencing both measurements;
- selection or measurement processes;
- coincidence within a small sample.

# Section 2: Pearson correlation

## Pearson's product-moment correlation

Pearson's correlation coefficient, usually written as $r$, measures the strength of a **linear** association.

- $r = 1$ represents a perfect positive linear relationship.
- $r = 0$ represents no linear relationship.
- $r = -1$ represents a perfect negative linear relationship.

```{r}
cor(
  iris$Sepal.Length,
  iris$Petal.Length,
  method = "pearson"
)
```

## Testing the association

```{r}
pearson_result <- cor.test(
  iris$Sepal.Length,
  iris$Petal.Length,
  method = "pearson"
)

pearson_result
```

The test evaluates the null hypothesis that the population correlation is zero. Report the coefficient, confidence interval, p-value, sample size, and scientific context.

## Interpreting magnitude

Descriptions such as weak, moderate, and strong depend on the subject area, measurement quality, and practical consequences. A modest correlation may matter greatly in one setting and have little practical relevance in another.

# Section 3: Assumptions and diagnostics

## Conditions for Pearson correlation

Pearson correlation is most informative when:

1. both variables are quantitative;
2. observations are independent;
3. the relationship is approximately linear;
4. influential outliers are absent or scientifically justified;
5. uncertainty is estimated under a suitable distributional model.

## Examine the shape

```{r}
plot(
  iris$Sepal.Length,
  iris$Petal.Length,
  pch = 19,
  col = rgb(0.1, 0.3, 0.7, 0.5),
  xlab = "Sepal length",
  ylab = "Petal length"
)
abline(lm(Petal.Length ~ Sepal.Length, data = iris), col = "firebrick", lwd = 2)
```

## Groups can change the story

```{r}
cols <- c("setosa" = "#1b9e77", "versicolor" = "#d95f02", "virginica" = "#7570b3")
plot(
  iris$Sepal.Length,
  iris$Petal.Length,
  pch = 19,
  col = cols[iris$Species],
  xlab = "Sepal length",
  ylab = "Petal length"
)
legend("topleft", legend = names(cols), col = cols, pch = 19, bty = "n")
```

The overall relationship may partly reflect differences between species, which illustrates why context and potential confounders belong in the interpretation.

# Section 4: Rank correlations and reporting

## Spearman's rank correlation

Spearman's coefficient measures a **monotonic** relationship using ranks. It is useful for ordinal variables, skewed measurements, or relationships that increase consistently without forming a straight line.

```{r}
spearman_result <- cor.test(
  iris$Sepal.Length,
  iris$Petal.Length,
  method = "spearman",
  exact = FALSE
)

spearman_result
```

## Kendall's tau

Kendall's tau is based on concordant and discordant pairs and is often useful with smaller samples or many tied ranks.

```{r}
kendall_result <- cor.test(
  iris$Sepal.Length,
  iris$Petal.Length,
  method = "kendall",
  exact = FALSE
)

kendall_result
```

## Choosing and reporting

Choose Pearson when the scientific question concerns a linear relationship and its assumptions are reasonable. Choose Spearman or Kendall when ranks and monotonic association better represent the variables and question.

A concise report should include:

- the variables and sample size;
- the selected coefficient and reason for choosing it;
- the estimated coefficient;
- uncertainty and p-value where appropriate;
- a graph and a contextual interpretation;
- a clear statement that association alone does not establish causation.
