---
title: "Non-Parametric Statistics"
subtitle: "C5025HF"
author: "Dr K.J. Mhango"
css: scroll.css
format:
  revealjs:
    theme: simple
    slide-number: true
    toc: true
    toc-depth: 3
    incremental: false
    code-line-numbers: false
    chalkboard: true
    code-fold: true
    code-tools: true
execute:
  echo: true
editor: visual
---

# Welcome

## What You Will Learn Today
- Parametric vs non-parametric intuition
- Ranking logic and distribution-free ideas
- Mann–Whitney, Wilcoxon, Kruskal–Wallis
- Correlations: Spearman & Kendall
- Exact tests (Fisher)


# 1: Why Non‑Parametrics?

## When parametric assumptions break
- Non-normal data
- Skewed distributions
- Ordinal measurements
- Heterogeneous variances
- Small samples
- Outliers

## Ranking-based thinking: The "Race" Analogy
Instead of looking at exact times (values), look at finishing positions (ranks).

### The Intuition: Team A vs Team B (Independent)
Imagine a race between two teams.

- **Parametric (t-test):** Compares the **Average Time** of Team A vs Team B.
  - If **one** runner in Team A takes 5 hours (outlier), Team A's *average* is ruined.
  
- **Non-Parametric (Mann-Whitney):** Compares the **Finishing Positions** (Ranks).
  - **It combines everyone** (both teams) into one single lineup from 1st to last.
  - Then it looks at where Team A falls vs Team B.
  - If that slow runner takes 5 hours or 5 days, they are still just **"Last Place"**. The rank (last) is the same.

### The Intuition: Before vs After (Paired)
Imagine the **same runner** running twice (Race 1 vs Race 2).

- **Parametric (Paired t-test):** Calculates the improvement (Time 1 - Time 2) for each runner, then averages them.
  - One runner improving by 5 hours makes the *whole group* look vastly improved on average.

- **Non-Parametric (Wilcoxon Signed-Rank):** 
  - It calculates the change for each runner.
  - **It ranks these changes by size** (ignoring direction). Small change = Rank 1, Huge change = Rank 100.
  - Then it asks: **"Are the biggest ranks associated with getting faster or getting slower?"**
  - If one runner improves by 5 hours, they get the highest rank, but they don't pull the *magnitude* of the sum to infinity. It caps the influence of the outlier.

### Why this is useful
- **Robust to Outliers:** Extreme values don't pull the results (unlike the mean).
- **Skew doesn't matter:** We are testing the order, not the bell curve.
- **Any units work:** cm, mm, log-transformed, or "Likert scales" — as long as you can say "A > B", you can rank.

### What the tests actually ask
- **Mann–Whitney (2 groups):** If I pick one random value from Group A and one from Group B, is A likely to be higher than B?
- **Wilcoxon (Paired):** It looks at the *change* in each pair. Do the **big changes** tend to be increases or decreases?
  - (It ranks the size of the changes, then checks if the positive ranks outweigh the negative ranks).
- **Kruskal–Wallis (>2 groups):** Do some groups systematically rank higher than others?

### What to report
- **Descriptive stats:** Medians and Interquartile Ranges (IQR) for each group.
- **Effect size / Direction:** Explain the *direction* of the difference in plain English.
  - Instead of just "p < 0.05", say: "Species A tends to have higher values than Species B."
  - Optional but good: "The probability that a random value from Group A exceeds Group B is approx X%."
- **Visuals:** Always include a boxplot or jitter plot to show the spread and overlap.

# 2: Choosing the Right Test

## Cheat Sheet
| Design | Parametric | Non-parametric |
|--------|------------|----------------|
| 2 independent groups | t-test | Mann–Whitney |
| 2 paired measures | paired t | Wilcoxon signed-rank |
| >2 independent groups | ANOVA | Kruskal–Wallis |
| Correlation | Pearson | Spearman, Kendall |
| 2×2 association | χ² test | Fisher exact |


# 3: t-test Using *iris*

## Goal
Compare **Sepal.Length** between two species.
We subset to two species for a t-test:
```{r}
iris2 <- subset(iris, Species %in% c("setosa", "versicolor"))
```

### Visualise
```{r}
#| fig-height: 3.5
library(ggplot2)
ggplot(iris2, aes(x = Species, y = Sepal.Length, fill = Species)) +
  geom_boxplot(alpha = 0.6,width=0.2) +
  geom_jitter(width = 0.1, alpha = 0.5) +
  theme_minimal() +
  ggtitle("Sepal Length by Species")
```

## Parametric t-test Assumptions
1. **Independence**
   - Observations must be independent.
   - In *iris*, each flower is a unique specimen → ✔ OK.

2. **Normality of the outcome within each group**
   - Test with Shapiro–Wilk.
```{r}
shapiro.test(iris2$Sepal.Length[iris2$Species=="setosa"])
shapiro.test(iris2$Sepal.Length[iris2$Species=="versicolor"])
```
 - If p < 0.05 in either group, normality is questionable for that group.
   - Also use Q–Q plot:
```{r}
library(ggplot2)
ggplot(iris2, aes(sample = Sepal.Length)) + stat_qq() + stat_qq_line() + facet_wrap(~Species)
```
 - Points close to the straight line suggest normality; strong bends or S‑shapes suggest non‑normality.

3. **Homogeneity of variances** (Levene test)
```{r}
library(car)
leveneTest(Sepal.Length ~ Species, data = iris2)
```
 - If p < 0.05, variances differ notably between the two species.

## What if assumptions fail?
- Non-normality → use **Mann–Whitney (Wilcoxon rank-sum)**.
- Heterogeneous variances → Welch t-test OR non-parametric.

## Parametric test
```{r}
t.test(Sepal.Length ~ Species, data = iris2)
```
 - Look for the p-value and the confidence interval; with normality and similar variances, this is appropriate.

## Non-parametric alternative (if assumptions break)
```{r}
wilcox.test(Sepal.Length ~ Species, data = iris2)
```
 - Use this when data are skewed or have outliers; the p-value tests if one group tends to have higher values.

### Example output and interpretation (R)
```{r}
# Run both to compare

tt_res <- t.test(Sepal.Length ~ Species, data = iris2)
mw_res <- wilcox.test(Sepal.Length ~ Species, data = iris2, exact = FALSE)

tt_res
mw_res

# Print key summaries for the slide
cat("t-test p-value:", format.pval(tt_res$p.value, digits = 3), "\n")
cat("Mann–Whitney p-value:", format.pval(mw_res$p.value, digits = 3), "\n")
```

- Interpret the p-values and medians.
- Interpretation: If data look non-normal or have outliers, prefer Mann–Whitney. Report simply, e.g., “Species A tends to have longer sepals than Species B.”

# Paired t-test Using *iris*

## Goal
Compare **Sepal.Length** vs **Sepal.Width** for the same flowers.
Since these measurements come from the *same* flower, they are paired.

### Visualise
```{r}
#| fig-height: 3.5
# Reshape data to long format for plotting
library(tidyr)
iris_long <- pivot_longer(iris, 
                          cols = c("Sepal.Length", "Sepal.Width"), 
                          names_to = "Measure", 
                          values_to = "Value")

ggplot(iris_long, aes(x = Measure, y = Value, fill = Measure)) +
  geom_boxplot(alpha = 0.6,width=0.2) +
  geom_jitter(width = 0.1, alpha = 0.2) +
  theme_minimal() +
  ggtitle("Comparison of Paired Measurements")
```


## Paired t-test Assumptions
1. **Independence of pairs**
   - Each flower is independent of others → ✔ OK.

2. **Normality of the differences**
   - We care about the distribution of $(Sepal.Length - Sepal.Width)$.
```{r}
diffs <- iris$Sepal.Length - iris$Sepal.Width
shapiro.test(diffs)
```
   - If p < 0.05, the differences are not normally distributed.
   - Check visually:
```{r}
#| fig-height: 4
ggplot(data.frame(diffs), aes(sample = diffs)) + stat_qq() + stat_qq_line() +
  ggtitle("Q-Q Plot of Differences")
```

## Parametric paired test
```{r}
t.test(iris$Sepal.Length, iris$Sepal.Width, paired = TRUE)
```
 - Checks if the mean difference is non-zero.

## Non-parametric alternative (if assumptions break)
### Wilcoxon Signed-Rank Test
```{r}
wilcox.test(iris$Sepal.Length, iris$Sepal.Width, paired = TRUE)
```
 - Tests if the median difference is non-zero (symmetric assumption) or if the distribution of differences is symmetric around zero.

### Example output and interpretation (R)
```{r}
pair_t <- t.test(iris$Sepal.Length, iris$Sepal.Width, paired = TRUE)
pair_w <- wilcox.test(iris$Sepal.Length, iris$Sepal.Width, paired = TRUE)

pair_t


pair_w

```
- Interpretation: “Sepal length is significantly different from sepal width within the same flowers.”

# Chi-square Test Using *iris*

## Goal
Test independence between **Species** and a discretised version of **Sepal.Width**.
We bin Sepal.Width:
```{r}
iris$WidthClass <- cut(iris$Sepal.Width, breaks = 3, labels = c("Narrow", "Medium", "Wide"))
tab <- table(iris$Species, iris$WidthClass)
```

### Visualise
```{r}
#| fig-height: 3.5
ggplot(iris, aes(x = Species, fill = WidthClass)) +
  geom_bar(position = "fill") +
  theme_minimal() +
  labs(y = "Proportion", title = "Width Class Distribution by Species") +
  scale_fill_brewer(palette = "Pastel1")
```

## Chi-square Assumptions
1. **Independence of observations** → ✔ OK.
2. **Expected cell counts ≥ 5** in most cells.
```{r}
chisq.test(tab)$expected
```
If expected counts are too small → chi-square invalid.
 - Scan the table: if any expected cell count is < 5, prefer Fisher’s Exact Test.

## Parametric-style categorical test
```{r}
chisq.test(tab)
```
 - Chi-square p-value tests association; warnings about small counts mean the test may be unreliable.

## Non-parametric alternative when assumptions fail
### Fisher's Exact Test
Works even with small expected cell counts.
```{r}
fisher.test(tab)
```
 - Fisher’s p-value is reliable even with small expected counts.

### Example output and interpretation (R)
```{r}
chisq_res <- suppressWarnings(chisq.test(tab))  # warning if small counts
fisher_res <- fisher.test(tab)

chisq_res
fisher_res

chisq_expected <- chisq_res$expected

chisq_expected

```

- Check expected counts; if any are < 5, prefer Fisher’s result.
- Interpretation: “Species and width class show evidence of association.”

# ANOVA Using *iris*

## Goal
Compare **Petal.Length** across all three species.
```{r}
iris$Species <- factor(iris$Species)
```

### Visualise
```{r}
#| fig-height: 3.5
ggplot(iris, aes(x = Species, y = Petal.Length, fill = Species)) +
  geom_boxplot(alpha = 0.6,width=0.2) +
  geom_jitter(width = 0.1, alpha = 0.5) +
  theme_minimal() +
  ggtitle("Petal Length by Species (3 Groups)")
```

## ANOVA Assumptions
1. **Independence** → ✔ by design.
2. **Normality within each group**
```{r}
by(iris$Petal.Length, iris$Species, shapiro.test)
```
 - If some species have p < 0.05, normality is doubtful for those groups.
3. **Homogeneity of variances** (Levene)
```{r}
leveneTest(Petal.Length ~ Species, data = iris)
```
 - If p < 0.05, variances differ across species.

If either normality or variance equality fails → cannot trust ANOVA F-test.

## Parametric ANOVA
```{r}
summary(aov(Petal.Length ~ Species, data = iris))
```
 - In the ANOVA table, check the row for Species: Pr(>F) is the p-value for group differences.

## Non-parametric alternative
### Kruskal–Wallis Test
```{r}
kruskal.test(Petal.Length ~ Species, data = iris)
```
 - Kruskal–Wallis p-value tests whether at least one group tends to have higher ranks than the others.

If significant → perform post-hoc Dunn tests.
```{r}
if (!requireNamespace("FSA", quietly = TRUE)) {
  cat("Package 'FSA' not installed; skipping Dunn post-hoc tests. Install with install.packages('FSA').\n")
} else {
  library(FSA)
  dunnTest(Petal.Length ~ Species, data = iris, method = "bonferroni")
}
```
 - Post-hoc Dunn tests (when available) show which pairs differ; use adjusted p-values to report significant pairs. If skipped, you can still report the Kruskal–Wallis result and show group medians/IQRs.

### Example output and interpretation (R)
```{r}
aov_res <- aov(Petal.Length ~ Species, data = iris)
kw_res  <- kruskal.test(Petal.Length ~ Species, data = iris)

summary(aov_res)
kw_res

# Extract and print concise p-values for the slide
s <- summary(aov_res)[[1]]
aov_p <- s[["Pr(>F)"]][1]

```

- Interpretation: With skew/outliers or non-normality, rely on Kruskal–Wallis. If significant, follow with pairwise comparisons (e.g., Dunn tests) and report which species differ.

# Correlation Using *iris*

## Goal
Assess the relationship between **Sepal.Length** and **Petal.Length**.

### Visualise
```{r}
#| fig-height: 3.5
ggplot(iris, aes(x = Sepal.Length, y = Petal.Length)) +
  geom_point(alpha = 0.6, color = "darkblue") +
  geom_smooth(method = "loess", color = "red", se = FALSE) +
  theme_minimal() +
  ggtitle("Sepal vs Petal Length")
```


## Parametric Correlation (Pearson) Assumptions
1. **Linearity**
   - The relationship should be a straight line.
2. **Normality**
   - Both variables should be normally distributed.
```{r}
shapiro.test(iris$Sepal.Length)
shapiro.test(iris$Petal.Length)
```
   - If p < 0.05, normality assumption is violated.

## Parametric test (Pearson)
```{r}
cor.test(iris$Sepal.Length, iris$Petal.Length, method = "pearson")
```
 - Tests for linear correlation.

## Non-parametric alternative (if assumptions break)
### Spearman's Rank Correlation
```{r}
cor.test(iris$Sepal.Length, iris$Petal.Length, method = "spearman")
```
 - Uses ranks; tests for *monotonic* relationship (doesn't have to be a straight line, just consistently increasing or decreasing).

### Kendall's Tau
```{r}
cor.test(iris$Sepal.Length, iris$Petal.Length, method = "kendall")
```
 - Better for small samples or many ties.


- Interpretation: "There is a strong positive monotonic correlation between sepal length and petal length."


# Summary of When to Use Non-Parametric Methods 


### Use a non-parametric test when:

- Data are **clearly non-normal** (Shapiro–Wilk fails; Q–Q plots bend strongly).
- **Variances differ strongly** across groups (Levene fails).
- Data are **ordinal** (ranks, Likert scores).
- There are **outliers** that distort means.
- Sample size is **small**, making CLT unreliable.
- Group distributions have **different shapes**, not just shifted means.
