---
title: "Data Visualisation Tutorial"
subtitle: "A self-guided Quarto tutorial"
author: "Dr Kanthu Joseph Mhango"
date: "2026-01-23"
format:
  html:
    toc: true
    toc-depth: 3
    number-sections: true
    code-fold: false
execute:
  echo: true
  warning: false
  message: false
editor: visual
---

## Welcome (read this first)

This tutorial assumes **you are at the beginning of your journey with R** and **you are seeing ggplot2 for the first time**. That is completely fine. We will be very slow, very explicit, and we will repeat the same ideas until they feel familiar.

### What “Base R plotting” means (the default plotting engine)

R comes with a built-in plotting system. It is often called **Base R graphics**.

-   It is the **default** because it is included with R.
-   You do **not** need to install anything to use it.
-   You usually build a plot by giving R **step-by-step drawing instructions**:
    -   “draw the points”
    -   “now draw a line”
    -   “now add a legend”
    -   “now change the axis labels”

This is called an **imperative** style: you tell the computer what to do next, like giving directions.

### What ggplot2 is (and why people use it)

`ggplot2` is a very popular R package for plots. It is based on an idea called the **Grammar of Graphics**.

The beginner-friendly meaning is:

-   A plot is not just a picture.
-   A plot is a **structured description** of how data becomes marks (points/lines/bars) on axes.

So instead of manually drawing each element, you *describe the meaning*:

-   which dataset
-   which variable goes on x and y
-   which variable controls colour (or size, or shape)
-   what mark to draw (points? lines? bars?)
-   whether to split into panels (facets)

This is called a **declarative** style: you declare what the plot means, and ggplot draws it.

### The big advantage of ggplot2 (consistency)

Base R can make good plots, but complex plots often require many manual steps.

ggplot2 is easier for many people because the code is **consistent**:

-   the beginning is almost always `ggplot(data, aes(...))`
-   each new visual layer looks like `+ geom_*()`
-   legends usually appear automatically from the mapping in `aes()`
-   multi-panel plots are usually one line: `facet_*()`

### The anatomy of ggplot code (learn this pattern)

Most ggplot graphs follow this repeatable “sentence”:

``` r
ggplot(DATA, aes(MAPPINGS)) +
  geom_SOMETHING(...) +
  scale_SOMETHING(...) +
  coord_SOMETHING(...) +
  facet_SOMETHING(...) +
  labs(...) +
  theme_SOMETHING(...)
```

Now in plain English:

-   `ggplot(DATA, ...)`: “I am making a plot from this dataset.”
-   `aes(...)`: “Here is what variables mean visually (x, y, colour, etc.).”
-   `geom_*()`: “Here is how to draw the data (points, lines, bars…).”
-   `scale_*()`: “Here is how to translate values into colours/axes/labels.”
-   `coord_*()`: “Here is the coordinate system / viewing window.”
-   `facet_*()`: “Split into multiple panels (small multiples).”
-   `labs(...)`: “Titles and axis labels so humans can read it.”
-   `theme_*()`: “Styling rules (font sizes, background, grid lines).”

### How this tutorial works (Base R → think → ggplot → discuss)

Every module follows the same rhythm:

1.  **We show the Base R version first** (manual drawing, step by step).
2.  **You pause and predict the ggplot version** (you try to guess the pieces).
3.  **We show the ggplot version** (using the same grammar pattern).
4.  **We discuss** the differences in very simple language.

## Setup

### Load packages and set defaults

```{r}
library(ggplot2)

set.seed(123)

# this below code is optional, it just makes plots consistent in this tutorial
knitr::opts_chunk$set(
  fig.width = 8,
  fig.height = 4.8,
  fig.align = "center"
)
```

### What we will use

-   Dataset: `iris`
-   Tools: **Base R** graphics and **ggplot2**

```{r}
head(iris)
```

::: callout-note
**A beginner-friendly way to think about good visualisation**

Every plot should answer a question. In this tutorial, we repeatedly ask:

-   What comparison or pattern should the reader see *in one glance*?
-   What is the **main visual signal** (x position, y position, colour, panels, text)?
-   What is the **most likely misread** (and how do we prevent it)?
:::

------------------------------------------------------------------------

## 1 — Scatter plot with legend + reference line

### Goal

Create a scatter plot of:

-   x = `Sepal.Length`
-   y = `Petal.Length`
-   colour by `Species`

Add:

-   a **dashed 1:1 reference line**
-   a **legend** with clear labels
-   axis labels with **units**

### Step 1: Base R version (shown first)

In Base R, you build a plot by telling R what to draw **one step at a time**.

Read the next code like a recipe:

-   `plot(...)` draws the points
-   `abline(...)` draws the dashed 1:1 line
-   `legend(...)` creates the legend

Here is the Base R code:

```{r}
plot(
  iris$Sepal.Length, iris$Petal.Length,
  col = as.numeric(iris$Species),
  pch = 16,
  xlab = "Sepal Length (cm)",
  ylab = "Petal Length (cm)",
  main = "Scatter with Dashed Reference Line"
)

abline(a = 0, b = 1, lty = 2)

legend(
  "topleft",
  legend = levels(iris$Species),
  col = 1:3,
  pch = 16,
  title = "Species"
)
```

::: callout-note
**Principles used**

-   **Reference lines are arguments**: a 1:1 line creates an immediate benchmark for “above vs below”.
-   **Figure–ground**: keep non-data ink light (dashed line) so points remain the focus.
-   **Legends must be decodable**: if colour encodes groups, the legend must be present and readable.
:::

### Step 2: pause and predict the ggplot2 version

Now we do the same figure in ggplot2. Before looking at the solution, pause and ask:

-   What is the dataset? (**iris**)
-   What goes on x? (**Sepal.Length**)
-   What goes on y? (**Petal.Length**)
-   What does colour mean? (**Species**)
-   What do we draw? (**points**)
-   What extra guide line do we add? (**a 1:1 line**)

Try writing a ggplot “sentence” in your head first. If you want to sketch code, use this practice chunk (it will not run):

```{r eval=FALSE}
# ggplot(iris, aes(x = Sepal.Length, y = Petal.Length, colour = Species)) +
#   geom_point(...) +
#   geom_abline(...) +
#   labs(...) +
#   theme_...()
```

### Step 3: ggplot2 version (solution)

```{r}
ggplot(iris, aes(Sepal.Length, Petal.Length, colour = Species)) +
  geom_point(size = 2, alpha = 0.85) +
  geom_abline(intercept = 0, slope = 1, linetype = 2) +
  labs(
    x = "Sepal Length (cm)",
    y = "Petal Length (cm)",
    colour = "Species",
    title = "Scatter with Dashed Reference Line"
  ) +
  theme_minimal(base_size = 13)
```

### Step 4: discussion (what you should notice)

-   In **Base R**, you built the plot by issuing separate commands.
-   In **ggplot2**, you declared the meaning once (inside `aes(...)`), and ggplot handled the legend automatically.
-   The ggplot version is easier to extend consistently: if you want another layer later, you add another line starting with `+`.

------------------------------------------------------------------------

## 2 — Grouped line plot

### Goal

Plot three lines (one per species) over an index `x = 1:5` using the provided means:

```{r}
x <- 1:5
setosa_means     <- c(1.4, 1.45, 1.5, 1.55, 1.6)
versicolor_means <- c(4.0, 4.1, 4.2, 4.3, 4.4)
virginica_means  <- c(5.5, 5.6, 5.7, 5.8, 5.9)
```

### Step 1: Base R version (shown first)

In Base R, a “grouped line plot” often means:

-   plot the first line with `plot(..., type = "l")`
-   then manually add other lines with `lines(...)`
-   then manually add a legend

Here is the Base R code:

```{r}
plot(
  x, setosa_means,
  type = "l", lwd = 2, col = 1,
  ylim = c(1, 6),
  xlab = "Index", ylab = "Mean Petal Length",
  main = "Grouped Line Plot by Species"
)
lines(x, versicolor_means, col = 2, lwd = 2)
lines(x, virginica_means,  col = 3, lwd = 2)
legend(
  "topleft",
  legend = c("setosa", "versicolor", "virginica"),
  col = 1:3, lwd = 2, title = "Species"
)
```

### Step 2: pause and predict the ggplot2 version

Now think in ggplot terms.

In ggplot2, you strongly prefer your data to be in a simple table where:

-   each row is one observation you want to draw
-   each column is a variable (x, y, group, etc.)

So, to draw three lines, we should build a table with:

-   `Index` (x)
-   `MeanPetalLength` (y)
-   `Species` (which line / which colour)

### Step 3: ggplot2 version (solution)

First we create a tidy dataframe:

```{r}
means_df <- data.frame(
  Index = rep(x, times = 3),
  Species = rep(c("setosa", "versicolor", "virginica"), each = length(x)),
  MeanPetalLength = c(setosa_means, versicolor_means, virginica_means)
)

head(means_df)
```

Now we plot it with the repeatable ggplot pattern:

```{r}
ggplot(means_df, aes(x = Index, y = MeanPetalLength, colour = Species)) +
  geom_line(linewidth = 1.2) +
  geom_point(size = 2) +
  labs(
    x = "Index",
    y = "Mean Petal Length",
    colour = "Species",
    title = "Grouped Line Plot by Species (ggplot2)"
  ) +
  theme_minimal(base_size = 13)
```

### Step 4: discussion (why ggplot feels easier here)

-   In **Base R**, you wrote separate code for each group (`setosa`, `versicolor`, `virginica`).
-   In **ggplot2**, groups are just **values in a column** (`Species`), so adding a new group would just mean adding rows to the data.
-   The legend is automatic again because we mapped `colour = Species`.

::: callout-note
**Principles used**

-   **Proximity + continuity**: line plots are best when the story is “change across ordered x”.
-   **Common scales**: fixed `ylim` (Base R) prevents misleading comparisons caused by autoscaling.
:::

------------------------------------------------------------------------

## 3 — Labelled plot (text marks)

### Goal

Replace points with initials:

-   setosa → “s”
-   versicolor → “v”
-   virginica → “v” (problem!)

This exposes a design issue: **labels can collide** and **symbols can be ambiguous**.

### Step 1: Base R version (shown first)

In Base R, a labelled plot usually works like this:

-   first you create a blank plotting area (`type = "n"`)
-   then you place text labels using `text(...)`

Here we label each point using the first letter of the species name.

::: callout-tip
### Solution (Base R)

```{r}
plot(
  iris$Sepal.Length, iris$Petal.Length,
  type = "n",
  xlab = "Sepal Length",
  ylab = "Petal Length",
  main = "Labelled Plot (Initials Instead of Points)"
)

text(
  iris$Sepal.Length, iris$Petal.Length,
  labels = substr(iris$Species, 1, 1),
  col = as.numeric(iris$Species)
)

legend(
  "topleft",
  legend = levels(iris$Species),
  col = 1:3, pch = 15, title = "Species"
)
```
:::

### Step 2: pause and predict the ggplot2 version

In ggplot2, the Base R function `text(...)` is replaced by a layer called **`geom_text()`**.

So the “translation” is:

-   Base R: “draw blank plot, then write text at coordinates”
-   ggplot2: “map x and y, map the label, then choose the text geometry”

Before looking at the solution, try to predict:

-   What should go in `aes(...)` if you want text labels? (hint: `label = ...`)
-   What geom draws text? (hint: `geom_text()`)

### Step 3: ggplot2 version (solution)

First, we create labels that are not ambiguous. “s/v/v” is confusing, so we will use **Se / Ve / Vi**.

```{r}
iris_lbl <- transform(
  iris,
  Abbr = c("Se", "Ve", "Vi")[as.numeric(Species)]
)

ggplot(iris_lbl, aes(x = Sepal.Length, y = Petal.Length, label = Abbr, colour = Species)) +
  geom_text(size = 3.2) +
  labs(
    x = "Sepal Length (cm)",
    y = "Petal Length (cm)",
    colour = "Species",
    title = "Labelled Plot with Text Marks (ggplot2)"
  ) +
  theme_minimal(base_size = 13)
```

### Step 4: discussion (what ggplot is doing for you)

-   In **Base R**, you had to remember a special trick: `type = "n"` to make a blank canvas, then `text(...)` to place labels.
-   In **ggplot2**, you use the same pattern as always:
    -   `ggplot(data, aes(...))` to declare meaning
    -   `+ geom_text()` to choose the mark
    -   `+ labs()` and `+ theme_*()` for readability
-   This consistency is a big reason ggplot is easier for beginners.

::: callout-warning
**Principle (avoid ambiguous encodings):** two classes share the same initial (“v”). If you use text marks, choose labels that are **unique** and **legible** (e.g., “Se”, “Ve”, “Vi”), or use colour + shape, or facets.
:::

------------------------------------------------------------------------

## 4 — Small multiples: Base R vs ggplot2 facets

### Goal

Create **three random subsets** of `iris` and show them side-by-side to compare patterns.

Key requirement:

-   all panels must share the **same axis limits** so comparisons are valid.

### Step 1: Base R version (shown first)

This is a classic situation where ggplot tends to feel easier: **small multiples** (many similar plots side-by-side).

In Base R you typically have to:

-   set the layout manually (`par(mfrow = ...)`)
-   use a loop to draw each panel
-   fix axis limits yourself so the panels are comparable

Here is the Base R code:

```{r}
set.seed(123)
n_subsets <- 3
subset_size <- 50

xlim <- range(iris$Sepal.Length)
ylim <- range(iris$Petal.Length)

par(mfrow = c(1, n_subsets), mar = c(4, 4, 3, 1))

for (i in 1:n_subsets) {
  sub <- iris[sample(nrow(iris), subset_size), ]
  plot(
    sub$Sepal.Length, sub$Petal.Length,
    col = as.numeric(sub$Species),
    pch = 16,
    xlim = xlim, ylim = ylim,
    xlab = "Sepal Length", ylab = "Petal Length",
    main = paste("Subset", i)
  )
}

par(mfrow = c(1, 1))
```

### Step 2: pause and predict the ggplot2 version

Now think in ggplot terms:

-   “If I had one table where every row is a point, and I add a column called `Subset`, then I can facet by that column.”

So your prediction should be:

-   create a column `Subset`
-   draw one scatter plot
-   ask ggplot to split into panels with `facet_wrap(~ Subset)`

### Step 3: ggplot2 version (solution)

```{r}
set.seed(123)
n_subsets <- 3
subset_size <- 50

subset_list <- lapply(1:n_subsets, function(i) {
  sub <- iris[sample(nrow(iris), subset_size), ]
  sub$Subset <- paste("Subset", i)
  sub
})
iris_subs <- do.call(rbind, subset_list)

ggplot(iris_subs, aes(Sepal.Length, Petal.Length, colour = Species)) +
  geom_point(size = 2) +
  scale_color_manual(values = c("darkgreen", "steelblue", "darkorange")) +
  coord_cartesian(
    xlim = range(iris$Sepal.Length),
    ylim = range(iris$Petal.Length)
  ) +
  facet_wrap(~ Subset) +
  labs(
    x = "Sepal Length (cm)",
    y = "Petal Length (cm)",
    colour = "Species",
    title = "Random Subsets of the Iris Dataset"
  ) +
  theme_minimal(base_size = 13) +
  theme(legend.position = "bottom")
```

### Step 4: discussion (why ggplot2 is “scalable” here)

-   In **Base R**, making 10 panels means more layout thinking, more looping, and more manual control.
-   In **ggplot2**, the plot code stays almost the same. You mostly change the **data** (more rows / more subsets), and ggplot keeps the layout consistent.
-   This is one of the main advantages of the Grammar of Graphics approach: once the mapping is declared, “many panels” becomes a data problem, not a drawing problem.

::: callout-note
**Principles used**

-   **Small multiples**: comparisons become easier when panels share scales and design.
-   **Consistency beats decoration**: repeatable structure reduces cognitive load.
-   **Grammar of Graphics**: once the mapping is declared, faceting is a data operation rather than a manual layout job.
:::

------------------------------------------------------------------------

## 5 — Scatter + smoother + scale thinking

### Goal

Add a group-wise smoother to `Sepal.Length` vs `Petal.Length`:

-   Base R: `lowess()` per species
-   ggplot2: `geom_smooth(method = "loess")`

Also practice **scale discipline**:

-   choose breaks intentionally
-   label them clearly

### Step 1: Base R version (shown first)

In Base R, adding a “trend line” per group usually means:

-   draw the scatter plot
-   split the data by group (Species)
-   for each group, compute a smoother (here we use `lowess(...)`)
-   draw that smoother line with `lines(...)`

Here is the Base R code:

```{r}
plot(
  iris$Sepal.Length, iris$Petal.Length,
  col = as.numeric(iris$Species),
  pch = 16,
  xlab = "Sepal Length",
  ylab = "Petal Length (cm)",
  main = "Sepal vs Petal Length with Mean Smoother"
)

for (i in 1:3) {
  sub <- iris[iris$Species == levels(iris$Species)[i], ]
  lines(lowess(sub$Sepal.Length, sub$Petal.Length), col = i, lwd = 2)
}

legend(
  "topleft",
  legend = levels(iris$Species),
  col = 1:3, pch = 16, lwd = 2,
  title = "Iris Species"
)
```

### Step 2: pause and predict the ggplot2 version

In ggplot2, the “translation” is usually:

-   points → `geom_point()`
-   smoother → `geom_smooth()`
-   grouping by Species happens automatically because colour is mapped to Species

So before you look, predict what the ggplot sentence will look like:

-   start: `ggplot(iris, aes(...))`
-   then: `+ geom_point(...)`
-   then: `+ geom_smooth(...)`
-   then: labels and theme

### Step 3: ggplot2 version (solution)

```{r}
ggplot(iris, aes(x = Sepal.Length, y = Petal.Length, colour = Species)) +
  geom_point(size = 2, alpha = 0.7) +
  geom_smooth(se = FALSE, method = "loess", linewidth = 1.2) +
  scale_color_manual(
    values = c("darkgreen", "steelblue", "darkorange"),
    name = "Iris Species"
  ) +
  labs(
    x = "Sepal Length (cm)",
    y = "Petal Length (cm)",
    title = "Sepal vs Petal Length with Mean Smoother"
  ) +
  theme_minimal(base_size = 13)
```

### Step 4: discussion (what is simpler in ggplot2)

-   In **Base R**, you manually looped over species and manually drew each smoother line.
-   In **ggplot2**, you told ggplot “colour means Species”, and ggplot uses that grouping to draw separate smoothers automatically.
-   This is the same “consistency advantage” again: you add layers, and ggplot handles grouping + legends in a predictable way.

::: callout-warning
**Principle (don’t let smoothers over-claim):** a smoother is not “truth”; it’s a modelling choice. Always ask:

-   Is the smoother appropriate for the data density and noise?
-   Would a linear model or summary by bins be clearer?
-   Should you show uncertainty (SE/CI ribbon) or keep it simple?
:::

------------------------------------------------------------------------

## Mini-section — Visualisation principles checklist

Before you submit a figure, check:

-   **Integrity**: are scales honest? any distortion? any unnecessary ink?
-   **Perception**:
    -   proximity: are related items grouped?
    -   similarity: are colours/shapes consistent?
    -   figure–ground: is the data visually dominant?
-   **Structure (Grammar of Graphics)**:
    -   are mappings (`aes`) explicit and consistent across plots?
    -   are scales and guides clearly labelled?
-   **Accessibility**:
    -   can it be read in grayscale?
    -   is colour the *only* encoding?
    -   is text large enough for projection/print?

------------------------------------------------------------------------

## Optional extensions

### Extension 1 — Make labels unambiguous

Replace initials with unique labels: “Se”, “Ve”, “Vi”.

### Extension 2 — Use facets to replace colour

Create a `facet_wrap(~ Species)` plot and move legend off (or drop colour entirely).

### Extension 3 — Write a “caption that completes the figure”

Write 3–4 sentences:

-   what is shown (variables + units)
-   what colour encodes
-   what the smoother means (if present)
-   what the key takeaway is (one sentence)

------------------------------------------------------------------------

## Wrap-up

You now have a repeatable workflow:

-   start with the **question**
-   choose encodings that support perception (Gestalt)
-   keep figures honest and minimal (Tufte)
-   build plots as structured statements (Grammar of Graphics)