---
title: "Measures of Central Tendency and Dispersion"
subtitle: "Deep foundations for statistical analysis and inference"
author: "K.J. Mhango"
css: scroll.css
format:
  revealjs:
    slide-number: true
    chalkboard: true
    toc: true
    transition: fade
    code-fold: false
    incremental: false
execute:
  echo: false
---

# Why do we summarise data?

A dataset can contain hundreds, thousands, or millions of values.

That creates a practical problem.

Even if all values are available, the human mind cannot easily grasp the overall pattern from a long list of raw numbers.

So statistics begins with compression.

We ask:

> Can we replace many numbers with a few meaningful summaries without destroying the truth in the data?

This is where measures of **central tendency** and **dispersion** come from.

---

# The danger of summary numbers

Suppose two groups both have a mean of 50.

That sounds like they are similar.

But one group might be tightly clustered around 50, while the other ranges from 10 to 90.

So a summary of the centre, by itself, is incomplete.

A single number can be useful.

A single number can also be deeply misleading.

---

# The two foundational questions

Most introductory descriptive statistics can be organised around two questions.

## Question 1: Where is the centre?

This leads to measures of **central tendency**.

## Question 2: How much do values vary around that centre?

This leads to measures of **dispersion**.

Together, these form the basic language of data description.

---

# Why this matters in statistical analysis and inference

These ideas are not just classroom definitions.

They sit underneath:

* regression models
* machine learning loss functions
* confidence intervals
* hypothesis tests
* experimental comparisons
* error analysis
* model evaluation

If you do not understand what centre and spread really mean, it becomes difficult to judge whether a result is stable, interpretable, or trustworthy.

---

# Part 1

# Central tendency

---

# R setup for examples

```{r}
#| echo: true
# Shared vectors and helper functions used in code examples
x <- c(2, 4, 6, 8)
y <- c(20, 21, 22, 23, 120)
z <- c(1.10, 1.20, 0.90)

get_mode <- function(v) {
  uniq <- unique(v)
  uniq[which.max(tabulate(match(v, uniq)))]
}

harmonic_mean <- function(v) {
  if (any(v == 0)) return(NA)
  length(v) / sum(1 / v)
}
```

---

# What is central tendency?

A measure of central tendency is a statistic intended to represent the **typical location** of the data.

But that phrase hides an important fact.

There is more than one sensible definition of “typical”.

For example:

* the balancing point of all values
* the middle-ranked observation
* the most common observation

These correspond to:

* mean
* median
* mode

Each one is answering a slightly different question.

---

# The arithmetic mean: idea first

The **arithmetic mean** is the quantity most people simply call “the average”.

Its deep idea is this:

> The mean is the balance point of the dataset.

Imagine all observations laid out on a number line, each with equal weight.

The mean is the point at which the dataset would balance if placed on a pivot.

This is why the mean is so central to mathematics and statistics: it behaves like a true centre of mass.

---

# The arithmetic mean: formula

$$
\bar{x} = \frac{1}{n}\sum_{i=1}^{n} x_i
$$

Let us unpack every part carefully.

* $x_i$ means the $i$-th observation in the dataset.
* $\sum_{i=1}^{n} x_i$ means add up all observations from the first to the last.
* $n$ is the number of observations.
* dividing by $n$ shares the total equally across all observations.

So the arithmetic mean is literally:

> total amount divided equally among all cases.

That is why it makes sense for quantities that combine by **addition**.

---

# Why “additive” data gives rise to the arithmetic mean

Suppose five fields yield:

2, 4, 6, 8, 10 t/ha

The total yield across the five observations is found by adding.

$$
2 + 4 + 6 + 8 + 10 = 30
$$

If we want the equal-share value per field, we divide by 5.

$$
30/5 = 6
$$

This is why the arithmetic mean is natural for quantities such as:

* weight
* height
* yield
* temperature measured on an appropriate scale
* test score
* residual error under squared-loss frameworks

The data combine by accumulation.

---

# Worked example: arithmetic mean

Dataset:

2, 4, 6, 8

Step 1: add the values

$$
2+4+6+8 = 20
$$

Step 2: count the values

$$
 n=4
$$

Step 3: divide total by count

$$
\bar{x}=20/4=5
$$

Notice that 5 is not one of the observed values.

The mean does not need to be an actual observation.

It is a mathematical centre, not necessarily a value someone recorded.

```{r}
#| echo: true
mean(x)
sum(x) / length(x)
```

---

# Important properties of the arithmetic mean

The arithmetic mean has several important mathematical properties.

## 1. It uses every observation

Every value contributes.

## 2. The deviations from the mean sum to zero

$$
\sum (x_i - \bar{x}) = 0
$$

This means the mean perfectly balances positive and negative deviations.

## 3. It minimises the sum of squared distances

Among all possible choices of centre, the mean is the value that makes

$$
\sum (x_i - c)^2
$$

as small as possible.

This is one reason why the mean appears in regression and machine learning.

---

# Weakness of the arithmetic mean

Because the mean uses every observation, it is sensitive to extreme values.

That can be a strength if extremes are real and meaningful.

It can also be a weakness if a few unusual values distort the story.

Example:

25, 27, 28, 30, 500

The mean is pulled upward strongly by 500.

So the mean is informative, but not always robust.

---

# Different kinds of means exist for a reason

A common question:

> Why are there several means? Why not just use one formula?

Because different quantities combine in different ways.

Three important cases are:

* additive processes → arithmetic mean
* multiplicative processes → geometric mean
* rates over equal quantities → harmonic mean

So the “right average” depends on the **structure of the process generating the data**.

> Why are there several means? Why not just use one formula?

---

# The geometric mean: intuition

The **geometric mean** is used when values combine **multiplicatively** rather than additively.

What does that mean?

It means the process evolves by repeated multiplication.

Examples:

* population multiplies by a growth factor each year
* investment multiplies by return factors over time
* bacterial growth multiplies
* fold-change measurements multiply relative to a baseline

If something changes by 1.10 one year and 0.90 the next, the total effect is not found by adding 1.10 and 0.90. It is found by multiplying them.

---

# Linear vs multiplicative change

A common confusion is thinking that **any quantity could be treated multiplicatively**. In principle that is true — but the key question is **how the process actually evolves**.

Some processes grow by **adding a constant amount each step**. Others grow by **multiplying by a factor each step**.

These two mechanisms produce very different trajectories.

| Step   | Linear (additive) growth | Multiplicative growth |
| ------ | ------------------------ | --------------------- |
| Start  | 100                      | 100                   |
| Step 1 | 100 + 20 = 120           | 100 × 1.20 = 120      |
| Step 2 | 120 + 20 = 140           | 120 × 1.20 = 144      |
| Step 3 | 140 + 20 = 160           | 144 × 1.20 = 172.8    |

In the **linear case**, the increase is always the same absolute amount.

In the **multiplicative case**, the increase depends on the current size of the quantity.

This difference explains why different averages are needed.

| Situation                      | Correct average |
| ------------------------------ | --------------- |
| Constant additive change       | Arithmetic mean |
| Constant multiplicative factor | Geometric mean  |

The geometric mean answers a specific question:

> If the quantity changed by a different multiplicative factor each step, what single constant factor would produce the same overall result?

For example, if a population evolves like this:

Start = 100

Year 1 factor = 1.20

Year 2 factor = 0.90

The total effect is

$$
100 \times 1.20 \times 0.90 = 108
$$

The geometric mean finds the **single growth factor that would reproduce this overall change if applied repeatedly**.

That is why the geometric mean is the natural summary when processes evolve through proportional change rather than constant addition.

---

# The geometric mean: formula

For positive values $(x_1, x_2, ..., x_n)$,

$$
GM = (x_1 x_2 \cdots x_n)^{1/n}
$$

Let us unpack this.

* Multiply all observations together.
* Then take the $n$-th root.

Why the $n$-th root?

Because multiplying all values together creates a total multiplicative product, and the $n$-th root asks:

> What constant factor, repeated $n$ times, would give the same overall product?

That is exactly the multiplicative analogue of the arithmetic mean.

---

# Why it is called “geometric”

Historically, the name comes from geometry and proportional reasoning.

The geometric mean appears naturally in problems involving ratios, proportions, and similar figures.

For two positive numbers $a$ and $b$, the geometric mean is:

$$
\sqrt{ab}
$$

This quantity arises in classical geometric constructions and proportional relationships.

So the name is historical and mathematical: it comes from the role of this mean in geometry and multiplicative proportion.

---

# Worked example: geometric mean

Suppose the growth factors over three periods are:

1.10, 1.20, 0.90

Then

$$
GM = (1.10 \times 1.20 \times 0.90)^{1/3}
$$

First multiply:

$$
1.10 \times 1.20 \times 0.90 = 1.188
$$

Now take the cube root:

$$
GM \approx 1.059
$$

So the typical multiplicative factor is about 1.059, meaning about 5.9% growth per period.

```{r}
#| echo: true
if (any(z <= 0)) stop("geometric mean needs positive values")
geom_mean <- exp(mean(log(z)))
geom_mean
```

---

# Why the arithmetic mean can fail for growth rates

Suppose returns are +50% and −50%.

The arithmetic mean of the percentages is 0%.

That sounds like no net change.

But starting from 100:

$$
100 \times 1.5 = 150
$$

then

$$
150 \times 0.5 = 75
$$

The final value is 75, not 100.

So arithmetic averaging misleads here because the process is multiplicative.

The geometric mean captures the compound effect correctly.

---

# Geometric mean and logs

A very useful identity is:

$$
GM = \exp\left(\frac{1}{n}\sum \ln(x_i)\right)
$$

This means:

1. take logs of the values
2. compute the arithmetic mean of the logs
3. transform back using the exponential

Why does this work?

Because logarithms turn multiplication into addition.

So the geometric mean is what you get when you move a multiplicative problem into an additive scale, average there, and then move back.

---

# Conditions and cautions for the geometric mean

The geometric mean requires positive values.

Why?

* you cannot take logs of zero or negative numbers in the usual real-valued sense
* multiplicative growth factors must normally be positive

So the geometric mean is not a universal replacement for the arithmetic mean.

It is appropriate when the underlying process genuinely behaves multiplicatively.

---

# The harmonic mean: idea first

The **harmonic mean** is used when averaging **rates** where a fixed amount of work, distance, or other denominator is being completed.

The classic example is speed.

Suppose you travel the same distance twice, but at different speeds each time.

The correct overall average speed is **not** the arithmetic mean of the two speeds.

Why not?

Because speed is a rate: distance per unit time.

What accumulates is not speed itself, but time.

---

# Harmonic mean and why rates are different

Suppose you travel:

* 60 miles at 30 mph
* 60 miles at 60 mph

Arithmetic mean of the speeds:

$$
(30+60)/2 = 45 \text{ mph}
$$

But check the actual times.

Time for first leg:

$$
60/30 = 2 \text{ hours}
$$

Time for second leg:

$$
60/60 = 1 \text{ hour}
$$

Total distance = 120 miles

Total time = 3 hours

True average speed:

$$
120/3 = 40 \text{ mph}
$$

So 45 mph is wrong.

---

# Harmonic mean: formula

$$
HM = \frac{n}{\sum_{i=1}^{n} 1/x_i}
$$

Let us unpack the logic.

* For a rate, the reciprocal ($1/x_i$) often represents time per unit quantity.
* We add these reciprocals because times accumulate additively.
* Then we invert again to return to the rate scale.

So the harmonic mean is the correct average when equal quantities are associated with different rates.

---

# Worked example: harmonic mean

For speeds 30 and 60 over equal distances:

$$
HM = \frac{2}{1/30 + 1/60}
$$

Compute the denominator:

$$
1/30 + 1/60 = 2/60 + 1/60 = 3/60 = 1/20
$$

Therefore:

$$
HM = \frac{2}{1/20} = 40
$$

This matches the true average speed.

```{r}
#| echo: true
harmonic_mean(c(30, 60))
```

---

# When the harmonic mean matters

The harmonic mean is useful for:

* average speed over equal distances
* average cost per unit under certain weighting structures
* ratios and rates in some scientific and computational settings
* combining precision-like quantities in some modelling contexts

The key question is always:

> Are I averaging raw values, multiplicative factors, or rates over equal quantities?

---

# Comparing arithmetic, geometric, and harmonic means

For positive data, these means are related by:

$$
AM \ge GM \ge HM
$$

Equality occurs when all observations are identical.

This is not just a curiosity.

It reflects how each mean responds differently to variation.

* the arithmetic mean is largest
* the harmonic mean is most pulled downward by small values
* the geometric mean lies in between

| **Mean Type**  | **Use when...**                                      | **Example**                                         |
| -------------- | ---------------------------------------------------- | --------------------------------------------------- |
| **Arithmetic** | The **Denominator** is constant (Equal Time).        | You run for 10 mins at 5mph, then 10 mins at 10mph. |
| **Harmonic**   | The **Numerator** is constant (Equal Distance/Work). | You run 1 mile at 5mph, then 1 mile at 10mph.       |
| **Geometric**  | The **Growth** is compounding (Multiplicative).      | Your investment grows 10% Year 1, then 20% Year 2.  |

---

# The median: idea first

The **median** is the middle value once the data are ordered.

This sounds simple, but it represents a profound change in perspective.

The mean uses the actual magnitudes of all values.

The median uses only their **order**.

So the median is not a balance point.

It is a **positional centre**.

---

# How to calculate the median

## Case 1: odd number of observations

Order the values and take the middle one.

Example:

3, 5, 7, 10, 12

There are 5 observations, so the middle position is the 3rd.

Median = 7

## Case 2: even number of observations

Order the values and average the two middle values.

Example:

3, 5, 7, 10

The two middle values are 5 and 7.

Median = ((5+7)/2 = 6)

```{r}
#| echo: true
median(x)
median(c(3, 5, 7, 10))
```

---

# Why the median makes sense

The median divides the ordered dataset into two equal halves.

At least half the observations lie at or below it.

At least half lie at or above it.

So the median answers:

> What value sits in the middle of the ranked data?

This is why it is often interpreted as a “typical” observation.

---

# Important properties of the median

## 1. It is resistant to outliers

Extreme values can move far away without changing the middle-ranked position much.

## 2. It works well for skewed distributions

Especially income, waiting times, and count data with long tails.

## 3. It is appropriate for ordinal data

If categories have an order but not equal intervals, the median may still make sense, while the mean may not.

## 4. It minimises absolute deviations

The median is the value that minimises:

$$
\sum |x_i - c|
$$

This mirrors the mean’s role for squared deviations.

---

# Mean versus median

This contrast is fundamental.

## The mean

* uses all magnitudes
* is mathematically elegant
* is sensitive to outliers
* minimises squared distance

## The median

* uses ordering, not full magnitude information
* is robust to outliers
* is ideal for skewed data
* minimises absolute distance

Neither is “always better”.

They answer different questions.

---

# Worked comparison: mean versus median

Dataset:

20, 21, 22, 23, 120

Mean:

$$
(20+21+22+23+120)/5 = 41.2
$$

Median:

Ordered data already shown, middle value = 22

The mean says the balance point is 41.2.

The median says the middle-ranked observation is 22.

If the goal is to describe a typical case, the median is often more faithful here.

---

# The mode: idea first

The **mode** is the most frequently occurring value.

This seems the simplest measure of centre, but it has a distinct interpretation.

The mode tells us:

> What value or category appears most often?

That makes it especially useful when frequency itself matters.

---

# How to calculate the mode

Count how often each value appears.

Example:

2, 4, 4, 5, 7, 7, 7, 9

Frequencies:

* 2 appears once
* 4 appears twice
* 5 appears once
* 7 appears three times
* 9 appears once

Mode = 7

```{r}
#| echo: true
get_mode(c(2, 4, 4, 5, 7, 7, 7, 9))
```

---

# Properties of the mode

## 1. It can be used for categorical data

This is very important.

For categories such as:

* favourite crop
* transport mode
* yes/no response
* disease class

there may be no meaningful mean or median, but the mode still makes sense.

## 2. A dataset can have more than one mode

* one mode → unimodal
* two modes → bimodal
* more than two → multimodal

## 3. It can be unstable in small samples

A small change in frequencies can change the mode.

## 4. It may not be unique

If several values tie for highest frequency, there may be several modes.

---

# Why multimodality matters

If a dataset is bimodal or multimodal, this often suggests that the data may contain **multiple underlying groups**.

For example:

* male and female body sizes mixed together
* two subpopulations in a biological experiment
* two customer segments in market data

So the mode is not just a descriptive statistic. It can hint at hidden structure.

---

# Which measure of centre should be used?

That depends on the data and the purpose.

## Use the mean when:

* values combine additively
* outliers are not too influential
* mathematical modelling requires it
* interval/ratio interpretation is meaningful

## Use the median when:

* data are skewed
* outliers are strong
* robustness matters
* ordinal scale data are involved

## Use the mode when:

* the question is about most common values or classes
* the data are categorical
* peaks in the distribution are substantively important

---

# Part 2

# Dispersion

---

# Why centre is not enough

Suppose both datasets have mean 10.

Dataset A:

10, 10, 10, 10, 10

Dataset B:

2, 6, 10, 14, 18

Same centre.

Very different variability.

So after asking “where is the centre?”, the next question is:

> How tightly or loosely are the data spread around that centre?

---

# What is dispersion?

Dispersion refers to the spread, variability, or scatter in the data.

A good measure of dispersion tells us how different observations are from one another, or from a chosen centre.

We will consider:

* range
* variance
* standard deviation
* coefficient of variation
* standard error
* confidence intervals

---

# Range

The **range** is the simplest measure of dispersion.

$$
Range = Max - Min
$$

Example:

3, 5, 7, 12

$$
Range = 12 - 3 = 9
$$

This tells us the total span of the data.

```{r}
#| echo: true
range(x)
diff(range(x))
```

---

# Strengths and weaknesses of range

## Strength

It is quick and intuitive.

## Weakness

It uses only two observations: the smallest and largest.

So it ignores almost everything in the dataset.

That means it is highly sensitive to outliers and unstable from sample to sample.

Useful as a rough first look, but weak as a serious description of variability.

---

# Variance: the big idea

Variance is one of the deepest ideas in statistics.

It asks:

> On average, how far are observations from the mean?

But there is a problem.

If we simply average raw deviations from the mean,

$$
\sum (x_i - \bar{x})
$$

we always get zero.

Positive and negative deviations cancel.

So we need a way to measure deviations without cancellation.

---

# Why we square deviations

Variance solves the cancellation problem by squaring the deviations.

$$
(x_i - \bar{x})^2
$$

Squaring does three things.

## 1. Makes all contributions non-negative

No cancellation.

## 2. Gives larger deviations more weight

Large departures from the mean matter more.

## 3. Produces a mathematically tractable quantity

This is one reason variance works so well in algebra, calculus, regression, and probability theory.

---

# Population variance

If we have every value in the population, the population variance is:

$$
\sigma^2 = \frac{1}{N}\sum_{i=1}^{N}(x_i-\mu)^2
$$

Here:

* $\mu$ is the population mean
* $N$ is the population size
* $\sigma^2$ is the population variance

Interpretation:

> the average squared deviation of all population values from the true population mean.

---

# Sample variance

In practice, we usually do not observe the whole population.

We observe a sample.

Then the sample variance is:

$$
s^2 = \frac{1}{n-1}\sum_{i=1}^{n}(x_i-\bar{x})^2
$$

This looks almost the same as population variance, but the denominator has changed.

Not $n$.

Instead, $n-1$.

That change is extremely important.

```{r}
#| echo: true
var(x)  # sample variance (n-1 denominator)
pop_var <- sum((x - mean(x))^2) / length(x)
pop_var
```

---

# Why not divide sample variance by n?

This is one of the most important questions in introductory statistics.

At first glance, dividing by $n$ seems natural, because there are $n$ observations.

But the sample mean ($\bar{x}$) is estimated from the same data.

That creates a subtle problem.

The deviations from $\bar{x}$ are not fully free to vary.

Once $n-1$ deviations are known, the last one is forced by the requirement that deviations sum to zero.

So only $n-1$ independent pieces of information remain.

That is the bridge to **degrees of freedom**.

---

# Q&A — why not apply an (n-1) correction to the mean?

**Q:** If estimating the mean uses the sample, why don't we "correct" the sample mean the way we correct variance (i.e., use an $n-1$ style adjustment)?

**A:** The sample mean $\bar{x}$ is an unbiased estimator of the population mean $\mu$ under random sampling: $E(\bar{x}) = \mu$. There is no systematic downward or upward bias to correct. Bessel's correction ($n-1$) is needed for variance because using $\bar{x}$ reduces the apparent variability in the sample — the sample deviations around $\bar{x}$ are constrained and would systematically underestimate the population variance if divided by $n$. That constraint does not create bias in the mean itself, so no analogous divisor is required for the mean.

---

# Bessel’s correction

Dividing by $(n-1)$ instead of $(n)$ is called **Bessel’s correction**.

It corrects the downward bias in sample variance.

That is why:

$$
s^2 = \frac{1}{n-1}\sum (x_i - \bar{x})^2
$$

is an **unbiased estimator** of the population variance under standard sampling assumptions.

In plain language:

> if you repeatedly sampled and calculated this quantity, its average would equal the true population variance.

---

# Worked example: sample variance

Dataset:

2, 4, 6, 8

## Step 1: compute the mean

$$
\bar{x} = (2+4+6+8)/4 = 5
$$

## Step 2: compute deviations from the mean

* 2 − 5 = −3
* 4 − 5 = −1
* 6 − 5 = 1
* 8 − 5 = 3

## Step 3: square the deviations

* 9
* 1
* 1
* 9

## Step 4: sum the squared deviations

$$
9+1+1+9 = 20
$$

$$
s^2 = 20/3 \approx 6.67
$$

```{r}
#| echo: true
data2 <- c(2, 4, 6, 8)
mean(data2)
devs <- data2 - mean(data2)
sum(devs^2)
var(data2)
```

---

# Interpreting variance

Variance is powerful, but its units are squared.

If the original data are in centimetres, the variance is in square centimetres.

If the data are in tonnes per hectare, variance is in squared tonnes per hectare.

That makes variance mathematically useful, but not always intuitive for direct interpretation.

This leads naturally to standard deviation.

---

# Standard deviation

The **standard deviation** is the square root of the variance.

$$
s = \sqrt{s^2}
$$

This returns the measure of spread to the original units of the data.

Using the previous example:

$$
s = \sqrt{6.67} \approx 2.58
$$

So the typical spread from the mean is about 2.58 units.

```{r}
#| echo: true
sd(x)
```

---

# Why standard deviation matters

Standard deviation is often easier to interpret than variance because it is in the same units as the data.

It is widely used because it connects to:

* normal distributions
* z-scores
* confidence intervals
* standard errors
* model residual analysis
* quality control

It is one of the central descriptive statistics in science and data analysis.

---

# A careful interpretation of standard deviation

A common interpretation is:

> Standard deviation is the average distance from the mean.

That is not exactly correct.

It is more accurate to say:

> Standard deviation is the square root of the average squared deviation from the mean.

That usually behaves somewhat like a typical distance, but it is not the same as the mean absolute deviation.

The distinction matters conceptually.

---

# Coefficient of variation

The **coefficient of variation** compares spread relative to the mean.

$$
CV = \frac{s}{\bar{x}}
$$

Often expressed as a percentage:

$$
CV\% = 100 \times \frac{s}{\bar{x}}
$$

This tells us how large the standard deviation is relative to the size of the mean.

```{r}
#| echo: true
cv <- sd(x) / mean(x)
cv_percent <- 100 * cv
cv
cv_percent
```

---

# Why coefficient of variation matters

Suppose:

Group A: mean = 100, SD = 10

Group B: mean = 10, SD = 10

The absolute SD is the same.

But the variability relative to the scale of the data is very different.

* A has CV = 10%
* B has CV = 100%

So CV is useful when comparing variability across datasets with different units or magnitudes, provided the scale makes ratio comparisons meaningful.

---

# Cautions for coefficient of variation

CV is not always appropriate.

It is most meaningful when:

* the data are on a ratio scale
* zero means an absence of the quantity
* the mean is positive and not close to zero

If the mean is close to zero, CV can become unstable or misleading.

---

# Standard error: a different concept

A common confusion is between **standard deviation** and **standard error**.

They are not the same thing.

## Standard deviation

describes variability among observed values.

## Standard error

describes uncertainty in an estimated statistic, usually the sample mean.

So standard deviation is about the **data**.

Standard error is about the **estimate**.

---

# Standard error of the mean

The standard error of the mean is:

$$
SE(\bar{x}) = \frac{s}{\sqrt{n}}
$$

Why this form?

Because averaging reduces noise.

As the sample size grows, the sample mean becomes more stable.

The denominator ($\sqrt{n}$) expresses that reduction in sampling variability.

```{r}
#| echo: true
se <- sd(x) / sqrt(length(x))
se
```

---

# Interpreting standard error

A small SE means the sample mean is estimated precisely.

A large SE means the sample mean is more uncertain.

Important:

A large sample can have:

* large SD

but still

* small SE

because even noisy data can estimate the mean precisely if enough observations are collected.

---

# Confidence intervals

A **confidence interval (CI)** provides a range of plausible values for a population parameter (commonly the mean), given the observed data and a chosen confidence level (e.g., 95%).

For the sample mean, a two-sided $(1-\alpha)100\%$ CI is:

$$
\bar{x} \pm t_{n-1, 1-\alpha/2} \times \frac{s}{\sqrt{n}}
$$

where $t_{n-1, 1-\alpha/2}$ is the critical value from the Student's $t$ distribution with $n-1$ degrees of freedom.

---

# Why confidence intervals matter

* They quantify uncertainty about parameters, not variability among observations.
* A 95% CI constructed by this method will, in repeated sampling, contain the true population mean approximately 95% of the time.
* The width of the CI depends on the sample variability ($s$), the sample size ($n$), and the chosen confidence level.

---

# Worked CI example

Suppose $\bar{x}=50$, $s=10$, $n=25$, and we want a 95% CI.

Degrees of freedom = 24.

The $t$ critical value (approx) $t_{24, 0.975}\approx 2.063$.

Standard error = $10/\sqrt{25} = 2$.

Margin of error = $2.063 \times 2 \approx 4.126$.

95% CI: $50 \pm 4.126 \Rightarrow (45.874, 54.126)$.

Interpretation: given the data and assumptions, this interval is a plausible range for the population mean.

```{r}
#| echo: true
t.test(x)$conf.int

alpha <- 0.05
n <- length(x)
se <- sd(x) / sqrt(n)
t_crit <- qt(1 - alpha / 2, df = n - 1)
me <- t_crit * se
mean(x) + c(-1, 1) * me
```

---

# CI for a difference in means (brief note)

When comparing two independent group means, construct a CI for the difference:

$$
(\bar{x}_1 - \bar{x}_2) \pm t_{df,1-\alpha/2} \times \sqrt{\frac{s_1^2}{n_1} + \frac{s_2^2}{n_2}}
$$

The degrees of freedom can be calculated with Welch's approximation when variances differ.

This CI replaces a sole focus on a point estimate and provides a sense of both magnitude and uncertainty for the difference.

```{r}
#| echo: true
x1 <- c(5, 6, 7, 8)
x2 <- c(2, 3, 4, 5)
t.test(x1, x2)$conf.int
```

---

# Why standard error and CI matter for inference

Standard error gives the building block (uncertainty of an estimate).

Confidence intervals place that uncertainty into a range and allow direct interpretation of plausible parameter values.

They are more informative than a single point estimate because they show both location and precision.

---

# Linking centre, spread, and inference

At this point, the structure should be clear.

## Measures of centre

summarise location.

## Measures of dispersion

summarise variability.

## Standard errors and confidence intervals

summarise uncertainty in estimated statistics.

These are related, but not interchangeable.

Confusing SD and SE often leads to misinterpretation of graphs, tables, and experimental results.

---

# Why these ideas are foundational in quantitative research

In quantitative research, these concepts appear constantly, even when hidden behind software.

Examples:

* feature scaling often uses means and standard deviations
* squared-error losses are built on mean and variance logic
* outlier detection depends on deviation from central tendency
* probabilistic models depend on mean–variance structure
* model residual diagnostics often examine spread and skew
* uncertainty estimates often depend on confidence intervals

Software can calculate these quantities instantly.

Understanding what they mean is the real skill.

---

# Common mistakes to avoid

## 1. Treating “average” as if it always means arithmetic mean

It does not.

## 2. Using the mean for heavily skewed data without reflection

## 3. Confusing SD with SE

## 4. Thinking variance is “just a formula” rather than a response to the cancellation problem

## 5. Using CV when the mean is near zero

---

# Summary table

| Measure            | Core idea                          | Best thought of as         |
| ------------------ | ---------------------------------- | -------------------------- |
| Arithmetic mean    | equal-share centre                 | additive average           |
| Geometric mean     | constant multiplicative factor     | compound average           |
| Harmonic mean      | average rate over equal quantities | reciprocal average         |
| Median             | middle-ranked value                | positional centre          |
| Mode               | most frequent value                | commonest value            |
| Range              | max minus min                      | total span                 |
| Variance           | average squared deviation          | mathematical spread        |
| Standard deviation | square root of variance            | spread in original units   |
| CV                 | SD relative to mean                | relative variability       |
| SE                 | SD of an estimate                  | uncertainty of a statistic |
| CI                 | plausible range for a parameter    | interval estimate          |

---

# Reflection questions

1. Why is the arithmetic mean natural for additive quantities?
2. Why is the geometric mean more suitable for growth factors?
3. Why does the harmonic mean appear naturally for rates?
4. Why can the median remain stable when the mean changes sharply?
5. Why does squaring solve the cancellation problem in variance?
6. Why does estimating the mean cost one degree of freedom?
7. Why is standard error about estimation rather than raw spread?

---

# Practice task 1: central tendency

For the dataset:

2, 3, 3, 4, 8, 20

Calculate and interpret:

* arithmetic mean
* median
* mode

Then explain which one best represents the typical value and why.

```{r}
#| echo: true
data1 <- c(2, 3, 3, 4, 8, 20)
mean(data1)
median(data1)
get_mode(data1)
```

---

# Practice task 2: choosing the right mean

A crop’s annual multiplicative growth factors are:

1.05, 1.10, 0.95, 1.08

1. calculate the arithmetic mean
2. calculate the geometric mean
3. explain which one is more appropriate and why

```{r}
#| echo: true
factors <- c(1.05, 1.10, 0.95, 1.08)
mean(factors)
exp(mean(log(factors)))
```

---

# Practice task 3: variance and standard deviation

```{r}
#| echo: true
data3 <- c(4, 5, 7, 8, 9)
mean(data3)
devs <- data3 - mean(data3)
sum(devs^2)
var(data3)
sd(data3)
```

For the dataset:

4, 5, 7, 8, 9

1. compute the sample mean
2. compute deviations from the mean
3. square and sum them
4. compute sample variance
5. compute sample standard deviation
6. interpret the result in words

---

# Final message

Descriptive statistics are not trivial preliminaries.

They are the conceptual foundation of quantitative reasoning.

If you understand what these measures are, how they are calculated, and why their formulas make sense, then later ideas in inference, modelling, and machine learning become far easier to understand.

If you only memorise the formulas, the rest of statistics will feel arbitrary.
