---
title: "Understanding the Chi-Square Test"
subtitle: "Counts, expected values, chi-square distributions, goodness-of-fit, independence, and assumptions"
author: "KJ Mhango"
format:
  revealjs:
    css: scroll.css
    theme: simple
    slide-number: true
    chalkboard: true
    transition: fade
    width: 1600
    height: 900
    margin: 0.08
execute:
  echo: true
  code-fold: true
  warning: false
  message: false
---

```{r setup}
#| include: false
set.seed(123)
library(ggplot2)
library(dplyr)
library(tidyr)
library(knitr)
```

## The problem chi-square solves

Many statistical tests are built for **measurements**.

Examples: height, yield, temperature, rainfall, body mass, exam score.

For these, means and differences between means make sense.

. . .

But many datasets are **counts in categories**.

Examples: pass/fail, healthy/diseased, species, location by species, treatment response.

---

## Categorical data need a different question

For categorical counts, the average category is usually not meaningful.

The mean of `healthy`, `infected`, and `dead` is not a biological quantity.

Instead, we ask:

> Are the counts we observed close to the counts we expected?

That is the central idea behind the chi-square test.

---

## The core intuition

The chi-square test is a test of **discrepancy**.

It compares:

**Observed counts**

against:

**Expected counts under a null hypothesis**

The test asks:

> If the null hypothesis were true, would this mismatch be ordinary or surprising?

---

## The hypothesis-testing logic

The chi-square test follows a familiar pattern.

1. State a null hypothesis.
2. Work out what the null hypothesis predicts.
3. Compare observed counts with expected counts.
4. Ask whether the mismatch is large relative to chance variation.

The data type is different from a t-test, but the logic is the same.

---

## A first example: rolling a die

Suppose we roll a six-sided die 60 times.

If the die is fair, each face has probability 1/6.

So the expected count for each face is:

```{r die-expected}
60 * (1/6)
```

So, under fairness, each face is expected about 10 times.

---

## Mild mismatch

| Face | Observed | Expected if fair |
|---:|---:|---:|
| 1 | 8 | 10 |
| 2 | 12 | 10 |
| 3 | 11 | 10 |
| 4 | 9 | 10 |
| 5 | 10 | 10 |
| 6 | 10 | 10 |

The observed counts are not exactly equal to expected counts.

That is normal. Random data rarely match expectation perfectly.

---

## Strong mismatch

| Face | Observed | Expected if fair |
|---:|---:|---:|
| 1 | 2 | 10 |
| 2 | 4 | 10 |
| 3 | 5 | 10 |
| 4 | 7 | 10 |
| 5 | 18 | 10 |
| 6 | 24 | 10 |

This looks more suspicious.

Faces 5 and 6 occur far more often than expected.

Faces 1, 2, and 3 occur far less often than expected.

---

## The chi-square statistic

Pearson's chi-square statistic is:

**χ² = sum of (Observed − Expected)² / Expected**

or more compactly:

**χ² = Σ (O − E)² / E**

where:

- **O** is an observed count
- **E** is an expected count
- **Σ** means add over categories or table cells

---

## Why observed minus expected?

**O − E** is the raw difference between what we saw and what the null hypothesis predicted.

If we expected 10 but observed 12, the difference is +2.

If we expected 10 but observed 8, the difference is −2.

Both are mismatches of size 2, but in opposite directions.

---

## Why square the difference?

The formula uses **(O − E)²**.

Squaring does two things.

1. It removes the sign.
2. It gives large mismatches more weight.

For example:

- 2² = 4
- 8² = 64

A difference of 8 contributes much more than a difference of 2.

---

## Why divide by expected?

The formula divides by **E**.

The same raw difference can mean different things at different scales.

A difference of 5 is huge when the expected count is 2.

A difference of 5 is tiny when the expected count is 500.

So chi-square measures mismatch **relative to expectation**.

---

## Scale example

If expected = 2 and observed = 7:

**(7 − 2)² / 2 = 12.5**

If expected = 500 and observed = 505:

**(505 − 500)² / 500 = 0.05**

Both have a raw difference of 5.

Only the first is strongly surprising relative to expectation.

---

## Worked example: mild die mismatch

```{r mild-die-table}
die <- tibble(
  face = factor(1:6),
  observed = c(8, 12, 11, 9, 10, 10),
  expected = rep(10, 6)
) |>
  mutate(
    difference = observed - expected,
    contribution = (observed - expected)^2 / expected
  )

kable(die, digits = 3)
```

---

## Mild die statistic

```{r mild-die-stat}
chi_die <- sum(die$contribution)
chi_die
```

This is small because observed counts are close to expected counts.

---

## Worked example: suspicious die

```{r suspicious-die-table}
suspicious_die <- tibble(
  face = factor(1:6),
  observed = c(2, 4, 5, 7, 18, 24),
  expected = rep(10, 6)
) |>
  mutate(
    difference = observed - expected,
    contribution = (observed - expected)^2 / expected
  )

kable(suspicious_die, digits = 3)
```

---

## Suspicious die statistic

```{r suspicious-die-stat}
chi_suspicious <- sum(suspicious_die$contribution)
chi_suspicious
```

This is much larger because the observed counts are far from expected counts.

---

## Observed versus expected

```{r die-plot, fig.width=10, fig.height=5}
bind_rows(
  die |> mutate(example = "Mild mismatch"),
  suspicious_die |> mutate(example = "Strong mismatch")
) |>
  pivot_longer(c(observed, expected), names_to = "type", values_to = "count") |>
  ggplot(aes(x = face, y = count, fill = type)) +
  geom_col(position = "dodge") +
  facet_wrap(~ example) +
  labs(
    title = "Observed counts compared with expected counts",
    x = "Die face",
    y = "Count",
    fill = "Count type"
  ) +
  theme_minimal(base_size = 18)
```

---

## From statistic to distribution

A chi-square statistic is always zero or positive.

It cannot be negative because it is built from squared differences.

The smallest possible value is 0.

That happens when every observed count exactly equals its expected count.

But a statistic alone is not enough.

We need to know whether the statistic is large **relative to chance**.

---

## The chi-square distribution

The chi-square distribution is the reference distribution for chi-square tests.

It answers:

> If the null hypothesis were true, what chi-square values would we expect from random sampling variation?

There is not just one chi-square curve.

There is a family of curves controlled by **degrees of freedom**.

---

## Degrees of freedom

Degrees of freedom describe how many values are free to vary after constraints have been imposed.

For a goodness-of-fit test with **k** categories:

**df = k − 1**

For a two-way table with **r** rows and **c** columns:

**df = (r − 1)(c − 1)**

---

## Chi-square distributions

```{r chi-square-distributions, fig.width=10, fig.height=5.5}
x <- seq(0, 25, length.out = 500)
chisq_curves <- expand.grid(
  x = x,
  df = c(1, 2, 5, 10)
) |>
  as_tibble() |>
  mutate(density = dchisq(x, df = df))

ggplot(chisq_curves, aes(x = x, y = density, colour = factor(df))) +
  geom_line(linewidth = 1.3) +
  labs(
    title = "Chi-square distributions with different degrees of freedom",
    x = "Chi-square value",
    y = "Density",
    colour = "df"
  ) +
  theme_minimal(base_size = 18)
```

---

## What degrees of freedom change

As degrees of freedom increase, the distribution shifts to the right and becomes more spread out.

That makes sense.

With more categories or more table cells, there are more opportunities for random discrepancies to accumulate.

So a chi-square value that is large for 1 df may not be large for 10 df.

---

## The p-value

Once we have a chi-square statistic and degrees of freedom, we calculate a p-value.

The p-value asks:

> If the null hypothesis were true, how often would we see a chi-square statistic at least as large as the one we observed?

A small p-value means the mismatch is unusual under the null hypothesis.

---

## P-value as a right-tail area

```{r p-value-plot, fig.width=10, fig.height=5.5}
observed_chi <- chi_suspicious
df_die <- 5
x <- seq(0, 40, length.out = 1000)
plot_data <- tibble(
  x = x,
  density = dchisq(x, df = df_die),
  tail = x >= observed_chi
)

ggplot(plot_data, aes(x = x, y = density)) +
  geom_area(data = filter(plot_data, tail), fill = "grey70") +
  geom_line(linewidth = 1.2) +
  geom_vline(xintercept = observed_chi, linetype = "dashed", linewidth = 1.2) +
  labs(
    title = "The p-value is the right-tail area",
    subtitle = paste0("Observed chi-square = ", round(observed_chi, 2), ", df = ", df_die),
    x = "Chi-square value",
    y = "Density"
  ) +
  theme_minimal(base_size = 18)
```

---

## A common p-value warning

The p-value is **not** the probability that the null hypothesis is true.

It is the probability of a statistic this extreme, or more extreme, **assuming the null hypothesis is true**.

The p-value is about the data under a model.

It is not the probability that the model is true.

---

## Test 1: goodness-of-fit

The chi-square goodness-of-fit test is used when we have:

- one categorical variable
- observed counts across categories
- a proposed expected distribution

It asks:

> Do the observed category counts fit the expected distribution?

---

## Goodness-of-fit hypotheses

Null hypothesis:

> The observed counts follow the expected distribution.

Alternative hypothesis:

> The observed counts do not follow the expected distribution.

For a fair die:

> Each face has probability 1/6.

The alternative says at least one face has a probability different from 1/6.

---

## Expected counts for goodness-of-fit

For category i:

**Expected count = total sample size × expected probability**

For a fair die with 60 rolls:

**Expected count = 60 × 1/6 = 10**

This is why each face has expected count 10.

---

## Goodness-of-fit in R

```{r chisq-goodness}
chisq.test(suspicious_die$observed, p = rep(1/6, 6))
```

This compares observed counts with the counts expected under a fair die.

A low p-value suggests the observed data are not consistent with the fair-die model.

---

## Interpreting goodness-of-fit

A significant goodness-of-fit test tells us:

> The observed distribution differs from the expected distribution.

It does not prove why.

Possible explanations include:

- the die is physically biased
- the rolling process is biased
- the data were recorded incorrectly
- the rolls were not independent

Statistics detects the mismatch. Science explains the mismatch.

---

## Test 2: independence

The chi-square test of independence is used when we have:

- two categorical variables
- counts in a contingency table

It asks:

> Are the two categorical variables associated, or are they independent?

---

## Example: species and location

```{r birds-table}
birds <- matrix(
  c(38, 12,
    24, 26,
    8, 42),
  nrow = 3,
  byrow = TRUE
)

rownames(birds) <- c("Glen Affric", "Glen Nevis", "Rannoch Moor")
colnames(birds) <- c("Golden Eagle", "Red Kite")

kable(birds)
```

Question:

> Are bird species and location independent?

---

## Independence hypotheses

Null hypothesis:

> The two categorical variables are independent.

Alternative hypothesis:

> The two categorical variables are associated.

For the bird example:

> Species and location are independent.

---

## What independence means

Independence means that knowing one variable does not change the distribution of the other.

For the bird example:

> The proportion of Golden Eagles and Red Kites should be roughly the same in each location.

If species proportions differ strongly by location, that suggests association.

---

## Expected counts for independence

For a cell in a two-way table:

**Expected count = row total × column total / grand total**

This gives the count expected if the row variable and column variable were independent.

Expected counts preserve the row totals and column totals.

---

## Expected counts in R

```{r birds-expected}
bird_test <- chisq.test(birds)
bird_test$expected |> round(2) |> kable()
```

These are the counts predicted by the independence model.

---

## Running the independence test

```{r bird-test}
bird_test
```

The chi-square statistic measures how far the observed table is from the expected table.

The p-value tells us how unusual that mismatch would be if species and location were independent.

---

## Which cells caused the mismatch?

```{r birds-contrib}
bird_contrib <- (bird_test$observed - bird_test$expected)^2 / bird_test$expected
round(bird_contrib, 2) |> kable()
```

The overall statistic is the sum of these cell contributions.

Large contributions show where the association is coming from.

---

## Observed versus expected table cells

```{r birds-plot, fig.width=10, fig.height=5.5}
birds_long <- as.data.frame(as.table(birds)) |>
  rename(location = Var1, species = Var2, observed = Freq)

expected_long <- as.data.frame(as.table(bird_test$expected)) |>
  rename(location = Var1, species = Var2, expected = Freq)

left_join(birds_long, expected_long, by = c("location", "species")) |>
  pivot_longer(c(observed, expected), names_to = "type", values_to = "count") |>
  ggplot(aes(x = location, y = count, fill = type)) +
  geom_col(position = "dodge") +
  facet_wrap(~ species) +
  labs(
    title = "Observed and expected bird counts under independence",
    x = "Location",
    y = "Count",
    fill = "Count type"
  ) +
  theme_minimal(base_size = 18) +
  theme(axis.text.x = element_text(angle = 25, hjust = 1))
```

---

## Goodness-of-fit versus independence

| Feature | Goodness-of-fit | Independence |
|---|---|---|
| Variables | One categorical variable | Two categorical variables |
| Data structure | Vector of counts | Contingency table |
| Question | Do counts match a proposed distribution? | Are variables associated? |
| Null | Distribution equals expected distribution | Variables are independent |
| Expected counts | From specified probabilities | From row and column totals |
| Degrees of freedom | k − 1 | (r − 1)(c − 1) |

---

## Simulation gives the intuition

The chi-square distribution can feel abstract.

Simulation makes it concrete.

Imagine the null hypothesis is true.

Then:

1. Generate many fake datasets under the null.
2. Calculate χ² for each fake dataset.
3. Ask how often fake datasets are as extreme as the observed dataset.

That is the p-value in action.

---

## Assumptions: why they matter

Chi-square tests are simple and useful.

But they rely on assumptions.

If the assumptions are badly violated, the p-value may be misleading.

The main assumptions are about:

- the type of data
- category structure
- independence of observations
- expected counts
- study design

---

## Assumption 1: data are counts

Chi-square tests are for **counts in categories**.

They are not for raw percentages, proportions, means, or continuous measurements.

This is suitable:

| Outcome | Count |
|---|---:|
| Healthy | 42 |
| Diseased | 18 |

A percentage alone loses sample size.

70% of 10 and 70% of 10,000 carry very different evidence.

---

## Assumption 2: categories are mutually exclusive

Each observation should belong to one category or one cell.

If categories are healthy, infected, and dead, one plant should not be counted in more than one category at the same time.

If observations can belong to multiple categories, the counts are not simple independent category counts.

---

## Assumption 3: observations are independent

One observation should not determine or strongly influence another observation.

Possible violations:

- many leaves sampled from the same plant
- many people sampled from the same household
- repeated observations from the same site
- clustered spatial observations
- paired before/after measurements

---

## When independence fails

If observations are clustered, paired, repeated, or spatially correlated, consider methods that represent the design.

Examples:

- logistic regression
- multinomial models
- mixed-effects models
- generalized linear mixed models
- permutation tests that respect the sampling structure

The right method depends on how the data were generated.

---

## Assumption 4: expected counts are large enough

The chi-square test uses an approximation.

The approximation works better when expected counts are not too small.

A common classroom rule is:

> Expected counts should generally be at least 5.

The important values are the **expected counts**, not just observed counts.

---

## Checking expected counts

```{r expected-count-check}
bird_test$expected |> round(2) |> kable()
```

If expected counts are too small, options include:

- collecting more data
- combining categories where scientifically sensible
- using Fisher's exact test for small 2 by 2 tables
- using simulation-based p-values
- using a model designed for sparse counts

---

## Assumption 5: categories should not be invented after seeing the result

If we repeatedly regroup categories until we find a significant result, the p-value becomes misleading.

Good practice:

- define categories before testing where possible
- distinguish confirmatory analysis from exploratory analysis
- report the analysis process honestly

---

## Assumption 6: sampling design must support interpretation

The chi-square test can detect a pattern.

It cannot repair a weak study design.

If locations were sampled in different seasons, by different observers, or with different methods, a significant location effect might partly reflect those design differences.

The test detects mismatch.

The study design determines whether the mismatch is interpretable.

---

## What chi-square does not tell you

A chi-square test does not automatically tell you practical importance.

With very large samples, tiny deviations can be statistically significant.

With small samples, large-looking deviations may not be statistically significant.

Always inspect observed counts, expected counts, percentages, cell contributions, and scientific context.

---

## Effect size: Cramer's V

For a contingency table, Cramer's V is one measure of association strength:

**V = sqrt(χ² / [n × min(r − 1, c − 1)])**

where:

- n is total sample size
- r is number of rows
- c is number of columns

---

## Cramer's V in R

```{r cramers-v}
chi <- unname(bird_test$statistic)
n <- sum(birds)
r <- nrow(birds)
c <- ncol(birds)
cramers_v <- sqrt(chi / (n * min(r - 1, c - 1)))
cramers_v
```

Cramer's V is 0 when there is no association.

Larger values indicate stronger association, but interpretation depends on context.

---

## Reporting goodness-of-fit

A concise report should include the test, χ² value, degrees of freedom, p-value, and interpretation.

```{r report-values-goodness, include=FALSE}
p_suspicious <- pchisq(chi_suspicious, df = 5, lower.tail = FALSE)
```

Example:

> A chi-square goodness-of-fit test suggested that the observed die rolls were not consistent with a fair die, χ²(df = 5) = `r round(chi_suspicious, 2)`, p = `r signif(p_suspicious, 3)`. Faces 5 and 6 occurred more often than expected under the fair-die model.

---

## Reporting independence

Example:

> A chi-square test of independence suggested an association between bird species and location, χ²(df = `r bird_test$parameter`) = `r round(bird_test$statistic, 2)`, p = `r signif(bird_test$p.value, 3)`. Inspection of observed and expected counts showed that Golden Eagles were observed more often than expected in Glen Affric and less often than expected in Rannoch Moor, while Red Kites showed the opposite pattern.

---

## Final summary

The chi-square test is used for categorical count data.

Its core idea is:

> Compare observed counts with expected counts.

The statistic is:

**χ² = Σ (O − E)² / E**

Large values mean large discrepancy.

The chi-square distribution tells us whether that discrepancy is unusual under the null hypothesis.

---

## Final summary: the two tests

Two common introductory chi-square tests are:

1. **Goodness-of-fit**  
   One categorical variable; asks whether observed counts match a proposed distribution.

2. **Independence**  
   Two categorical variables; asks whether the variables are associated.

In both cases, the test is only as meaningful as the null hypothesis, expected counts, assumptions, and study design.

---
